# Generalized Linear Models

In this chapter, we discuss a family of models called *generalized linear models*. These models include ordinary least squares regression and many others. All[^3] of the models presented in this chapter can be realized as examples of this common framework. We won't dwell on the common framework in this book, but focus on two specific examples - *logistic regression* and *Poisson regression*. We will also be in the classical framework of regression, where we are interested in the theory and assessing the fit of the model, rather than focusing on predictive modeling. We will do *a lot* of estimates of regression coefficients for various models in this chapter. I hope that by the time you get to the end of reading it, you will be able to set these up in your sleep.

Since generalized linear models include OLS, let's take a brief review of OLS. Recall that our model is

\[
y_i = \beta_0 + \sum_{j = 1}^p \beta_j x_{ij} + \epsilon_i
\]

where $\epsilon$ is normal with mean 0 and unknown variance $\sigma^2$. Another way of thinking about this is that we are modeling the *expected value* of the response via a linear (affine) equation.

\[
E[y_i] = \beta_0 + \sum_{j = 1}^p \beta_j x_{ij}
\]

and we are assuming that we have *normal errors* or deviations from the expected value. Yet another way to think about it is that the response $y_i$ is normal with mean $\mu = \beta_0 + \sum_{i = j}^p \beta_j x_{ij}$ and variance $\sigma^2$.

## Logistic Regression

We begin with an example. 

### Malaria Example (Intro)

Consider the `malaria` data set in the `fosdata` package. Mice were bit 10 times by mosquitoes, and the number of salivary gland sporozoites detected in the mosquitoes was measured. The mice were then observed to determine whether they got malaria. Some of the mice were given a malarial antibody pre-mosquito bite. Our goals are to determine whether the number of sporozoites affects the probability of infection, and whether the antibody affects the probability of infection.

```{r}
library(tidyverse)
mm <- fosdata::malaria |> 
  as_tibble()
ggplot(mm, aes(x = factor(malaria), y = sporozoite)) +
  geom_boxplot() +
  geom_jitter(width = .1, height = .1) 
```

One question we should have is whether mosquitos with 0 sporozoites should be included in the study at all, because they can't possibly infect the mice. If we remove those mosquitos, we get the following graph.

```{r}
mm <- filter(mm, sporozoite > 0)
ggplot(mm, aes(x = factor(malaria), y = sporozoite)) +
  geom_boxplot() +
  geom_jitter(width = .1, height = .1)
```

We want to think of the number of sporozoites as the *predictor* though, which is usually on the $x$ axis.

```{r}
ggplot(mm, aes(x = sporozoite, y = malaria)) +
  geom_point() + 
  geom_smooth(method = "lm")
```

Here we see an issue with this approach: if the line is supposed to represent the probability of getting malaria, then it is impossible for the line to be above 1 or below 0, but the line does go above 1. This is a problem with using OLS to model a binary response, though it is still sometimes approrpriate in certain situations. We will use *logistic regression*.

```{r}
ggplot(mm, aes(x = sporozoite, y = malaria)) +
  geom_point() + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"))
```

We can see with one observed sporozoite, the probability of getting malaria is about 25%, while with 20 or more, the probability is essentially 100%. 

```{r}
ggplot(mm, aes(x = sporozoite, y = malaria)) +
  geom_point() + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"))
```

Now let's add the `antibody` variable and see whether the graphs look different.

```{r}
ggplot(mm, aes(x = sporozoite, y = malaria, color = factor(antibody))) +
  geom_jitter(width = .2, height = 0, alpha = .5) + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"))
```

Based on this plot, it seems that having the antibody protects from mosquitos which have a low viral load, but does not help against mosquitos with a high viral load.

### The model (theory)


Logistic regression is used to model responses that can only take two values. Such variables are called *binary* variables. Examples include determining whether a flipped coin comes up as Heads or Tails, whether a person recovers or doesn't recover from a disease, or whether a person votes yes or no in an election. Many times, we have additional information that is naturally grouped with the outcome. For example, we may have a person's age, height, weight, the amount and type of medicine they were given in addition to whether they recovered or didn't recover from the disease. The goal is to be able to model the recovery or non-recovery from the disease on the predictors.

Let's call the response variable $Y$, and assume for now that we have a single continuous predictor. The response would correspond to whether the mouse got malaria in the previous example, and the predictor would be the number of sporozoites. In OLS, we model 

\[
y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

so that the **expected value of the response** is modeled by $\beta_0 + \beta_1 x_i$. The $\epsilon_i$ is the error term. Reasoning similarly, we might reasonably model the expected value of the binary response as $\beta_0 + \beta_1 x_i$. Let's think carefully about what the expected value of a 0/1 binary random variable represents. Let $p$ be the probability of success. Then,

\[
E[Y] = 1 * p + 0 * (1 - p) = p
\]

So, modeling the expected value of the response is the same as modeling the probability of the response being 1. Then, our model would be (without the error term)

\[
p_i = \beta_0 + \beta_1 x_i
\]

This is fine, but we note that for some values of the parameters and predictors, we could get probabilities that are larger than 1 or smaller than 0. This would be difficult to interpret. So, it is common to model the **log-odds** of the probability instead.

\[
\log \frac{p}{1 - p} = \beta_0 + \beta_1 x_i
\]

The odds $\frac{p}{p - 1}$ can be any positive number and the log-odds can be any real number. Given the log-odds $L$ of an event, we can recompute the probability of the event via $p = \frac{e^L}{1 + E^L}$.

**Aside** Interpreting log-odds. The log-odds of an event is the log of the odds of the event. The odds of an event is the probability of the event divided by the probability of the event not happening. So, if we have a 50% chance of an event happening, then the odds are 1, and the log-odds are 0. If we have a 75% chance of an event happening, then the odds are 3, and the log-odds are $\log(3)$. If we have a 25% chance of an event happening, then the odds are $\frac{1}{3}$, and the log-odds are $\log(\frac{1}{3})$. 

If the log-odds increases by 1, then the odds increase by a multiplicative factor of $e$. If the log-odds increases by $\beta_1$, then the odds increase by a factor of $e^{\beta_1}$. So, if $\beta_1$ is positive, then the odds of the event happening increase as $x$ increases. If $\beta_1$ is negative, then the odds of the event happening decrease as $x$ increases. If $\beta_1$ is zero, then the odds of the event happening do not change as $x$ increases.

In the OLS model, we explicitly had an error term $\epsilon_i$. In the logistic regression model, we don't have an explicit error term, but we can think of the response as being a random variable that is generated according to the probabilities given by the model. That is, we can think of $Y_i$ as being a Bernoulli random variable with parameter $p_i$, where $p_i$ is given by the logistic function of $\beta_0 + \beta_1 x_i$.

### Estimating parameters using MLE

Let's see how to use MLE and the `optim` function to estimate the parameters $\beta_0$ and $\beta_1$ in the above model. We use the `fosdata::malaria` data set for specificity. 

We start by noting that the **likelihood** of a 0/1 random variable is determined by the probability of success. If $Y = 1$, then the likelihood of the data point is $p$ (probability of success) and if $Y = 0$, then the likelihood of the data point is $1 - p$ (probability of failure). We can write this succinctly as $p^Y (1 - p)^{1 - Y}$.

Therefore, the likelihood of observing a sequence of 0/1 random variables $Y_1, \ldots, Y_n$ with probabilities of success $p_1, \ldots, p_n$ is given by $\Pi_{i = 1}^n p_i^{Y_i} (1 - p_i)^{1 - Y_i}$, where $p_i$ is the probability of success for observation $i$. The log of the likelihood function simplifies to $\sum_{i = 1}^n Y_i \log(p_i) + (1 - Y_i) \log(1 - p_i)$.

Finally, we have the relationship between $p_i$ and the predictors and parameters: $p_i = \frac{1}{1 + e^{-\beta_0 - \beta_1 x_i}}$. Therefore, we can write the log-likelihood function in terms of $\beta_0$ and $\beta_1$ as follows:

```{r}
mm <- fosdata::malaria
nll <- function(par) {
  beta0 <- par[1]
  beta1 <- par[2]
  p <- boot::inv.logit(beta0 + beta1 * mm$sporozoite)
  -1 * sum(mm$malaria * log(p) + (1 - mm$malaria) * log(1 - p))
}
optim(par = c(0,0), fn = nll)
```

Let's compare those to what we would get using the built-in R function `glm`, which is the workhorse for doing all things with fixed effect generalized linear models.

```{r}
glm(malaria ~ sporozoite, data = mm, family = "binomial")
```

### Estimating parameters via loss function

Just like in OLS, there are alternative ways to estimate the parameters associated with logistic regression. In OLS, we saw that minimizing the SSE gave the same parameter estimates as maximizing the likelihood function. For logistic regression, the error function we want to minimize is:

\[
\frac{1}n \sum_{i = 1}^n \Phi(\eta_i y_i)
\]

where $\Phi = \log(1 + e^{-x})$ and $y_i$ has been recoded so that 1 is success and $-1$ is failure, and $\eta_i = \log {p_i}{1 - p_i}$ is the log-odds of $p_i = \beta_0 + \beta_1 x_i$.  Let's think about why this is a function that **penalizes** being away from the true observed values. The graph of $\Phi$ looks like this

```{r}
curve(log(1 + exp(-x)), from = -3, to = 3)
```

So this penalty penalizes **negative** values of $\eta_i y_i$ and encourages **positive** values of $\eta_i y_i$. If $y_i$ is 1, then we want $\eta_i$ to be positive, so that $\eta_i y_i$ is positive. If $y_i$ is $-1$, then we want $\eta_i$ to be negative, so that $\eta_i y_i$ is positive. Therefore, this penalty encourages the model to predict positive values for the predictors when the response is 1 and negative values for the predictors when the response is $-1$. Recall that large positive values of the log-odds correspond to high probabilities of success, and large negative values of the log-odds correspond to low probabilities of success. So, this penalty encourages the model to predict high probabilities of success when the response is 1 and low probabilities of success when the response is $-1$. 

What isn't at all obvious is that minimizing this function will yield the same parameter estimates as maximizing the likelihood function, but let's see it in an example.

```{r}
pen <- function(par) {
  beta0 <- par[1]
  beta1 <- par[2]
  p <- boot::inv.logit(beta0 + beta1 * mm$sporozoite)
  y <- ifelse(mm$malaria == 1, 1, -1)
  mean(log(1 + exp(-1 * (log(p/(1-p))) * y)))
}
optim(par = c(0,0), fn = pen)
```

### Interpreting summary

Now, we turn to interpreting the summary of `glm`. 

```{r}
mod <- glm(malaria ~ sporozoite, data = mm, family = "binomial")
summary(mod)
```

We have an estimate for intercept of -1.44 and for $\beta_1$ of 0.33341. The intercept tells us the log-odds of getting malaria when the number of sporozoites is 0 (according to the model - we actually know this is zero). The log-odds of getting malaria when the number of sporozoites is 0 is -1.44, which corresponds to a probability of getting malaria of $\frac{e^{-1.44}}{1 + e^{-1.44}}$. In R, this is computed via the `boot::inv.logit(-1.44)` function, which gives us a probability of about 0.19. So, according to the model, the probability of getting malaria when the number of sporozoites is 0 is about 19%.

The sporozoite estimate tells us that the log-odds of getting malaria increases by 0.33341 for every one unit increase in the number of sporozoites. Therefore, the odds of getting malaria will increase by a multiplicative factor of $e^{0.33341}$ for every one unit increase in the number of sporozoites. In R, this is computed via `exp(0.33341)`, which gives us a multiplicative factor of about 1.4. So, according to the model, the odds of getting malaria will increase by a factor of about 40 percent for every one unit increase in the number of sporozoites. 

The standard error, $z$ value and $p$ value have meanings similar to what we covered in OLS. The deviance plays a similar role to the sum of squared errors in OLS, and the null deviance is the deviance of the model with only an intercept. The difference between the null deviance and the residual deviance is a measure of how much better the model with the predictor is compared to the model with only an intercept. Formally, the null deviance is defined as -2 times the log likelihood of the model with only an intercept, and the residual deviance is defined as -2 times the log likelihood of the model with the predictor. Therefore, the difference between the null deviance and the residual deviance is 2 times the difference in log likelihoods between the two models. This can be used to perform a likelihood ratio test to determine whether the model with the predictor is significantly better than the model with only an intercept. Since we have already computed the likelihood, we can quickly confirm the deviance computations.

```{r}
#use nll from above - note that we already had a negative sign there
2 * nll(c(-1.44, 0.33341))
```

This should be the same as the residual deviance in the summary output, which it is!

We can compute an analogy of $R^2$ for logistic regression, which is called McFadden's $R^2$. This is defined as $1 - \frac{\text{residual deviance}}{\text{null deviance}}$. In R, this is computed as follows:

```{r}
1 - (mod$deviance / mod$null.deviance)
#using the built in function from the package pscl
library(pscl)
pR2(mod)
```

We can also perform a hypothesis test on which of two models is better. For example, we can compare the model with the predictor `sporozoite` to the model with `sporozoite` and `antibody`. 

```{r}
mod2 <- glm(malaria ~ sporozoite + antibody, data = mm, family = "binomial")
summary(mod2)
anova(mod, mod2, test = "Chisq")
```

Note that there is **not** statistically significant evidence that the model with `antibody` is better than the model with just `sporozoite`. We could also have looked directly at the $p$-values in `summary(mod2)`:

```{r}
summary(mod2)
```

We get slightly different $p$-values because the tests use different assumptions. One utility of the test using deviances is that it can be used to test the null hypothesis that a group of parameters is all zero versus the alternative that at least one of them is non-zero, which is not something that we can do with the $p$-values in `summary(mod2)`. For the test using deviances to be valid, the models **must be nested**. If the models are not nested, then it is common to use AIC to decide between the models, rather than a formal hypothesis test.

```{r}
AIC(mod, mod2)
```

Note that AIC chooses the model **with** antibodies while deviances eliminate antibodies. This goes back to the purpose of the two techniques - there are some theoretical reasons to prefer AIC for predictive modeling, and using $p$-values is sometimes prefered for building parsimonious or explanatory models. For completeness, we could also include the interaction between `sporozoites` and `antibody`:

```{r}
mod3 <- glm(malaria ~ sporozoite * antibody, data = mm, family = "binomial")
summary(mod3)
```

The model coefficients are `(Intercept)`, which is the log-odds of getting malaria when the mosquito has zero sporozoites and the mouse does not have the antibody, `sporozoite`, which is the change in log-odds of getting malaria for every one unit increase in the number of sporozoites when the mouse does not have the antibody, `antibody`, which is the change in log-odds of getting malaria for having the antibody when the number of sporozoites is **zero**, and `sporozoite:antibody`, which is the change in log-odds of getting malaria for every one unit increase in the number of sporozoites when the mouse has the antibody compared to when it does not have the antibody. 

**Note**: it can be hard to directly see whether the positive `antibody` is associated with **more** malaria or **less** malaria in the model with interaction. For example, if the typical mosquito has 50 sporozoites, then the effect of the antibody on the log-odds of getting malaria would be $-1.38 + (0.34 + .37) * 50 - 4.8$, compared to $-1.38 + .34 * 50$ for the antibody negative mice. The difference in these is $.37 * 50 - 4.8 = 13.7$, which means our model would say it is **much** more likely to get malaria with the antibodies. 

We can also use the `predict` function to get point predictions and confidence intervals for the probability of getting malaria associated with sporozoites and antibodies. We return to the model with `sporozoite` and `antibody`. You can read about the options in `predict` by typing `predict.glm`. Note that `predict` is a generic function which has different behavior depending on whether the object passed to it was fit using `lm` or `glm`. The one we had been using was `predict.lm`. 

```{r}
predict(mod2, newdata = data.frame(sporozoite = 10, antibody = "yes")) #log-odds
predict(mod2, newdata = data.frame(sporozoite = 10, antibody = "yes"), type = "response") #probability
predict(mod2, newdata = data.frame(sporozoite = 10, antibody = "no"), type = "response", se.fit = T) 
```

Note that we would want to interpret the standard error with caution here. First of all, it is the value that corresponds to **confidence** intervals in OLS, not **prediction** intervals. Second, it would be non-standard to compute a two-sided confidence interval for $p$ by taking $0.8868761 \pm 2 * 0.03624307$. A better way to compute a confidence interval would be to compute the symmetric interval in the log-odds scale, and convert it into the probability scale.

```{r}
predict(mod2, newdata = data.frame(sporozoite = 10, antibody = "no"), se.fit = T) 
boot::inv.logit(2.059222 + 2 * 0.3612499) #upper bound
boot::inv.logit(2.059222 - 2 * 0.3612499) #lower bound
```

When communicating results from logistic regression, it can be challenging to explain your results. The `marginaleffects` package has methods which allow for easier interpretation. First, we look at the **average marginal effect**. The average marginal effect of sporozoite estimates the *derivative* of the probability of getting malaria at each of the 139 data points and averages them. This gives an estimate for the *average effect* that a one unit change in sporozoite will have; where average means averging over the observed data.

```{r}
library(marginaleffects)
summary(mod2)
marginaleffects::avg_slopes(mod2)
```

The `avg_slopes` function estimates the derivative or the probability with respect to a continuous predictor, or the change in probability per category change for categorical predictors. This is often what you want to do, but in the case of `sporozoite`, the predictor is actually count data, so the derivative is hard to interpret. The `avg_comparisons` function estimates the change in probability for a one unit change in a continuous predictor, so for the case of sporozoites, it would estimate the average increase in probability assocaited with a one sporozoite increase.

```{r}
marginaleffects::avg_comparisons(mod2)
```

Sometimes, it is desirable to get the effect of a different unit of change.

```{r}
iqr_spor <- round(IQR(mm$sporozoite))
iqr_spor
avg_comparisons(mod2, variables = list(sporozoite = iqr_spor), type = "response")
```


On average, the probability of getting malaria increases by about 4.44 percentage points per increase in sporozoite. An increase of 6 spororzoites is associated with a 18.5 percentage point increase in the probability of getting malaria, averaged across conditions in the data set. Of course, neither of these statements is true when there are already 25 sporozoites, as the probability is already essentiall 1. Similarly, the average effect of not having the antibody is a 19 percentage point increase in the probability of getting malaria. We can rederive the result of `avg_comparisons` (without confidence bands):

```{r}
p1 <- predict(mod2, type = "response")
p2 <- predict(mod2, newdata = mm |> mutate(sporozoite = sporozoite + 1), type = "response")
mean(p2 - p1)
```

We can also directly visualize the probabilities.

```{r}
marginaleffects::plot_slopes(mod2, variables = "sporozoite", by = "antibody")
plot_predictions(mod2, condition = "sporozoite")
```

This gives the probabilities at each sporozoite level; the other values are set at their mean or mode. In the case of `sporozoite`, the default is that the antibody is set to `no`. 


```{r}
avg_predictions(mod2, newdata = datagrid(sporozoite = 1)) #sets other variables to typical values
predict(mod2, newdata = data.frame(sporozoite = 1, antibody = "no"), type = "response")

avg_predictions(mod2, newdata = datagrid(sporozoite = 1, grid_type = "counterfactual")) #averages over the values observed in study
mean(predict(mod2, newdata = mm |> mutate(sporozoite = 1), type = "response"))
```

The word *marginal* generally means to average over the other values. If we want to average the probabilities over the values observed in the data, then we need to specify.

```{r}
plot_predictions(mod2, by = "sporozoite", newdata = datagrid(sporozoite = seq(0, 30, by = 0.5), grid_type = "counterfactual"))
```

## Mixed Effects Logistic Regression

Let's consider the `Contraception` data set in the `mlmRev` package. The goal is to model contraceptive usage among women in Bangladesh. This is an *exploratory* analysis.

```{r}
cc <- mlmRev::Contraception |> 
  janitor::clean_names() |> 
  as_tibble()
summary(cc)
```

There are 60 different districts, and we aren't interested in modeling each one separately, but rather in modeling the spread among the districts using `district` as a random effect.

```{r}
mod1 <- lme4::glmer(use ~ livch + age + urban + (1|district), data = cc, family = "binomial")
summary(mod1)
```


```{r}
mod2 <- lme4::glmer(use ~ livch + age + urban + (1 + urban|district), data = cc, family = "binomial")
summary(mod2)
```

Let's think carefully about what the difference in the two models is. In the first model, we are allowing the intercept to vary across districts, but we are assuming that the effect of `urban` is the same across all districts. In the second model, we are allowing both the intercept and the effect of `urban` to vary across districts. The second model is more flexible, but it also has more parameters, so we would want to use AIC or a likelihood ratio test to determine which model is better. 

Looking at the coefficients, it appears that there may only be a difference between *having* a child and *not* having a child as it pertains to contraceptive use. Let's see how we can test that.

```{r}
cc$anychild <- factor(ifelse(cc$livch == "0", "none", "any"))

mod3 <- lme4::glmer(use ~ anychild + age + urban + (1 + urban | district),
                    data = cc, family = "binomial")

summary(mod3)
anova(mod2, mod3, test = "Chisq")
```

Indeed, the effect is based on whether there is a child at all; not how many children. Finally, let's examine the `age` variable. It seems unlikely that contraceptive use depends only on age; it seems plausible that usage would increase up to a certain age, and then start to decrease again. Let's add a quadratic term and see what we get.

```{r}
cc$agec <- cc$age - mean(cc$age)
cc$agec <- cc$agec/sd(cc$agec)
mod4 <- lme4::glmer(use ~ anychild + agec + I(agec^2) + urban + (1 + urban | district),
                    data = cc, family = "binomial")
summary(mod4)
anova(mod3, mod4)
```

The quadratic term is significant. We can compute the marginal effect of age by averaging over the other observed values in the data set, as below. When doing this, you need to find the approximate largest and smallest value of the `by` variable in your data, and create a sequence of values from the minimum to the maximumm.

```{r}
library(marginaleffects)
summary(cc$agec)
plot_predictions(mod4, by = "agec", type = "response", newdata = datagrid(agec = seq(-1.5, 2.2, by = 0.1), grid_type = "counterfactual"))
```

Alternatively, you could plot the estimates contraceptive use by age for typical values of the other covariates. Typical in this case means either mean for numeric variables or mode for categorical data, and the random effect is considered to be zero. In this instance, there are only small differences in the two plots, but as a general rule for mixed effect models, I would default to the marginal effect above.

```{r}
plot_predictions(mod4, condition = "agec", type = "response")
```

To get estimates that are more interpretable than the log-odds estimates, we can again use `marginaleffects`. Note that the marginal effect of `age` is not easily interpretable, because of the quadratic nature of that relationship.

```{r}
avg_slopes(mod4)
```

This indicates that having a child is associated with a 0.17687 increase in the probability scale; that is, an 18 percentage point increase in contraceptive use averaged across all data points, and urban areas are associated with a 16 percentage point increase in contraceptive use, averaging over the other covariates in the study.

Now that we have seen an example, let's write out the model for mixed effect logistic regression explicitly. Suppose we have one continuous predictor $x$, then the random intercepts model would be 

\[
\log \frac{p_{ij}}{1 - p_{ij}} = \beta_0 + \beta_1 x_{ij} + u_j
\]

where $p_{ij}$ is the probability of success for observation $i$ in group $j$, $\beta_0$ is the fixed intercept, $\beta_1$ is the fixed effect of the predictor, and $u_j$ is the random effect for group $j$. The random effects are assumed to be normally distributed with mean 0 and variance $\sigma^2_u$. The error term is not explicitly included in this model, but we can think of the response as being a Bernoulli random variable with parameter $p_{ij}$, where $p_{ij}$ is given by the logistic function of $\beta_0 + \beta_1 x_{ij} + u_j$.

### Nested random effects

In the data set `ardata::beethoven_tempos`, we had two categorical variables `symphony` and `movement`. Each symphony consisted of 4 movements, numbered 1-4. So, the data point `movement==1` could mean different things for each symphony. We day that `movement` is **nested** in `symphony`. If we want to correctly identify this in a mixed effect model, the random effect should be written as `(1|symphony/movement)`.  This ensures that R understands that `movement` is **nested** inside of `symphony`. Let's look at the `guImmun` data in the `mlmRev` package.

```{r}
gg <- mlmRev::guImmun |> 
  janitor::clean_names() |> 
  as_tibble()
summary(gg)
```

In this case, we see that we have two possible random effects: community and mom, with mom nested inside of community. If each mom is given a unique identifier, then we are OK, but to be safe, we can indicate the random effect of `mom` and `comm` via `(1|comm/mom)`. This is good practice because it indicates that we know that the random effects are nested, even if it is not necessary for a given data set.

## Count Regression

In this section, we suppose that the response variable is a **count** of some object. For example, the number of times a person goes to the doctor in a year, the number of times a person goes to the gym in a week, or the number of times a person eats out in a month. A common model for count data is the Poisson regression model. In this model, we assume that the response variable $Y$ is a Poisson random variable with mean parameter $\lambda$, where $\lambda$ is given by the exponential function of a linear combination of the predictors. Since counts can only be nonnegative numbers, the expected number of occurences will be  positive and we model

\[
log E[Y] = log \lambda = \beta_0 + \beta_1 x_1 + \ldots + \beta_P x_P
\]

The error term is not explicitly included in this model, but we can think of the response as being a Poisson random variable with parameter $\lambda$, where $\lambda$ is given by the exponential function of a linear combination of the predictors. 

We illustrate this by looking at the `bioChemists` data set in the `pscl` package. We wish to model `art`, the count of articles produced during the last 3 years of the Ph.D. In this example, we only model it on `ment`, which is the count of articles produced by Ph.D. mentor during the last 3 years, which we treat as numeric.

The model is 

\[
\log \lambda_i = \beta_0 + \beta_1 x_i
\]

whre $\lambda_i$ is the expected number of articles produced by the $i$th Ph.D. student, $\beta_0$ is the expected number of articles produced by a Ph.D. student whose mentor has produced 0 articles, and $\beta_1$ is the change in the expected number of articles produced by the Ph.D. student for every one article increase in the number of articles produced by the mentor. 

The **likelihood** of the data given values of $\beta_0$ and $\beta_1$ is 

\[
\Pi_{i = 1}^n \mathtext{dpois}(y_i, \beta_0 + \beta_1 x_i)
\]
or taking logs, it is `sum(dpois(y, betao + beta1 x, log = T))`. This is the function we want to maximize for our MLE.

```{r}
bb <- pscl::bioChemists
nll <- function(par) {
  beta0 <- par[1]
  beta1 <- par[2]
  -1 * sum(dpois(bb$art, exp(beta0 + beta1 * bb$ment), log = T))
}
optim(par = c(0,0), fn = nll)
```

We get an estimate of 0.26 for $\beta_0$ and 0.026 for $\beta_1$. Let's compare with `glm`.

```{r}
glm(art ~ ment, data = bb, family = poisson)
```

Let's now model the number of articles carefully on the other variables.

```{r}
mod <- glm(art ~ fem + mar + kid5 + phd + ment, data = bb, family = poisson)
summary(mod)
```

Recall that our model says that the **expected value** of the response is given by an affine combination of the predictors, and the **observed value** of the response is a sample from a Poisson random variable with that expected value. Poisson random variables have the property that their *variance* is equal to their *mean*. This will be one of our two primary diagnostics when seeing whether we have appropriately modeled a count variable.

To intuitively test whether the expected value is equal to the variance, we can compute all of the expected values

```{r}
evals <- predict(mod, type = "response")
```

and then use these to estimate the variance:

```{r}
mean(evals)
mean((mean(evals) - bb$art)^2)
```

We see that we have a **rough** estimate of 3.7 for the variance and 1.7 for the mean. This indicates that the variance might be larger than the mean, in which case we say the response is **overdispersed** or exhibits **overdispersion**.

Here are three more principled approaches to estimating the dispersion:

```{r}
mean((evals - bb$art)^2/evals)
mod$deviance/mod$df.residual #this is very common in R, because you can read it directly off of summary(mod)
performance::check_overdispersion(mod)
```

However you slice it, it appears that the data is not well-modeled by Poisson regression. One option for proceeding is to use *quasipoisson*, and this is often done in base R.

```{r}
mod <- glm(art ~ fem + mar + kid5 + phd + ment, data = bb, family = quasipoisson)
summary(mod)
```

The second common issue in count regression is *zero inflation*. That is when there are more zero values than would be expected from the model. The quick way to check that is"

```{r}
performance::check_zeroinflation(mod)
```

We can see that there are more observed zeros than predicted zeros, and we are likely underfitting the number of zeros. At this point, we want to fix that, and the cleanest way it so move away from base R and from `lme4` to use the `glmmTMB` package. Let's start over.

```{r}
library(glmmTMB)
mod <- glmmTMB::glmmTMB(art ~ fem + mar + kid5 + phd + ment, data = bb, family = poisson)
summary(mod)
performance::check_overdispersion(mod) 
```

OK, we see overdispersion. We will change our random model of the response from *Poisson* to *negative binomial*. Recall negative binomial random variables model the number of failures until a certain number of successes, where the probability of success is determined by the predictors. This is a more flexible model for count data, which allows the variance to be greater than the mean.
 
The base R command for computing probabilities related to negative binomial rvs is `dnbinom`, `pnbinom`, `qnbinom`, and `rnbinom`. 
The parameters are `size` and `prob`, where `size` is the number of successes and `prob` is the probability of success. 
The mean of a negative binomial rv is `size * (1 - prob) / prob` and the variance is `size * (1 - prob) / prob^2`.
 
The formula for the probability of x failures before the nth success with a probability of success p is (x + n - 1)!/(n - 1)!x! p^n (1 - p)^x

Replacing y! with Gamma(y + 1), we can extend this definition to the number of successes being **noninteger**. There is no real interpretation for this,
but it turns out to be useful for modeling.
 
We use the parameterization with size = dispersion and mu = mean. In this parameterization, dispersion \to infinity corresponds to the Poisson distribution.

This is NOT the same notion of dispersion used in count regression (e.g., quasi-Poisson), where Var(Y) = phi * mu and phi = 1 corresponds to Poisson.

The conflict arises because negative binomial dispersion controls a quadratic variance term (mu + mu^2 / dispersion), whereas quasi-Poisson dispersion is a multiplicative scaling of mu.
 
At any rate, the crucial fact is that by choosing the size parameter carefully we can have thevariance of the response vary with the size of the response. A natural guess is that the variance should
 be the mean^2. If $E[Y] = 1$ then $aY$ has mean a and variance a^2, for example. I am not saying that
 there is any kind of multiplicative thing going on here, but variance being proportional to the mean^2
 makes some sense. 
 
 We can also try the variance being proportional to the mean itself (rather than **equal** to the mean itself). That is
 also implemented in `glmmTMB`.
 
 Let's see how all of this works.
 
```{r}
mod2 <- glmmTMB::glmmTMB(art ~ fem + mar + kid5 + phd + ment, data = bb, family = nbinom2) #assumes variance = mu + mu^2/dispersion
summary(mod2)
performance::check_overdispersion(mod2) #looks good
performance::check_zeroinflation(mod2)
```
 
We notice that switching to a negative binomial random model fixed both the overdispersion and the zero inflation! That's really nice; it looks like our data is pretty well modeled by negative binomial. We also could have tried:

```{r}
mod3 <- glmmTMB::glmmTMB(art ~ fem + mar + kid5 + phd + ment, data = bb, family = nbinom1) #assumes variance = mu + mu/dispersion
AIC(mod2, mod3)
```

We see that the model 2 is preferred. Now we can trust the $p$-values a little bit better, so we can do some variable selection to get a parsimonious model. We remove `mar` and `phd` to get the following model.

```{r}
mod4 <- glmmTMB::glmmTMB(art ~ fem + kid5 + ment, data = bb, family = nbinom2) 
summary(mod4)
performance::check_overdispersion(mod4)
performance::check_zeroinflation(mod4)
```

As before, we can use `marginaleffects` to get easier interpretations of the coefficients.

```{r}
library(marginaleffects)
avg_comparisons(mod4)
summary(bb)
plot_predictions(mod4, by = "ment", newdata = datagrid(ment = seq(0, 77, by = 1), grid_type = "counterfactual"), vcov = TRUE)
```

### Mixed Effects Model

Consider the `Owls` data set in the `glmmTMB` package. The ChatGTP summary of the accompanying article is: Baby barn owls use loud calls to ask their parents for food, and these calls can show how hungry they are. The study found that chicks beg more loudly when their mother is present than when their father is there, even though they are just as hungry in both cases. Researchers think this happens because mothers and fathers act differently when feeding, so chicks adjust their behavior to get more food. This shows that young birds can tell their parents apart and change how they ask for food to improve their chances of being fed. 

```{r}
library(glmmTMB)
oo <- glmmTMB::Owls |> 
  janitor::clean_names() |> 
  as_tibble()
oo
```

We start by modeling `sibling_negotiation` by the other variables.

```{r}
mod1 <- glmmTMB::glmmTMB(sibling_negotiation ~ sex_parent + arrival_time + brood_size + food_treatment + (1|nest), data = oo, family = poisson)
summary(mod1)
performance::check_overdispersion(mod1)
```

We see that the model has significant overdispersion, so we switch to negative binomial.

```{r}
mod2 <- glmmTMB::glmmTMB(sibling_negotiation ~ sex_parent + arrival_time + brood_size + food_treatment + (1|nest), data = oo, family = nbinom2)
summary(mod2)
performance::check_overdispersion(mod2)
performance::check_zeroinflation(mod2)
```

This indicates that the model is not yet correctly specified. Let's fix the zero inflation and see whether that also helps the **underdispersion**.

```{r}
mod3 <- glmmTMB::glmmTMB(sibling_negotiation ~ sex_parent + arrival_time + brood_size + food_treatment + (1|nest), 
                         ziformula = ~1, 
                         data = oo, 
                         family = nbinom2)
summary(mod3)
performance::check_overdispersion(mod3)
performance::check_zeroinflation(mod3)
```

Note that fixing the zero inflation also fixed the dispersion issues. So, this looks like it is pretty good. We note that there is no sex of parent effect on the number of calls. That is because sibling negotiation measures the number of calls that the siblings make **before** the parent arrives. We can also test to see whether there is an interaction between food treatment and sex of the parent that arrives.

```{r}
mod4 <- glmmTMB::glmmTMB(sibling_negotiation ~ sex_parent * food_treatment + arrival_time + brood_size + (1|nest), 
                         ziformula = ~1, 
                         data = oo, 
                         family = nbinom2)
summary(mod4)
```

There is not a significant interaction. Question: what would an interaction have meant in terms of the owls? What would have been a plausible explanation for an interaction, depending on the exact nature of the study?

## Offsets

There is one additional applied regression tool that we have not yet talked about - *offsets*. An *offset* is used when the response variable is a count, but the counts are not all measured over the same amount of time. For example, if we are modeling the number of doctor visits in a year, but some people are only observed for 6 months, then we would want to use an offset to account for the fact that some people are observed for a shorter amount of time. The offset is typically the log of the amount of time that each observation is observed for. This allows us to model the rate of doctor visits per unit of time, rather than the total number of doctor visits. Mathematically, we suppose that the expected count is given by 

\[
E[\mathrm{count}_i] = \lambda_i t_i
\]
where $\lambda_i$ is the rate at which the counts appear and $t_i$ is the length of time observed. Then we can model the log of the expected value as 
\[
\log E[\mathrm{count}_i] = \log \lambda_i + \log t_i = \beta_0 + \beta_1 x_{1i} + \log t_i
\]
where we are using the familiar $\log \lambda_i = \beta_0 + \beta_1 x_{1i}$ formulation when we have exactly one predictor. The theory specifies that the coefficient associated with $\log t_i$ will be exactly 1, even if a different coefficient has a higher likelihood or a lower loss function. In order to specify that the coefficient is exactly 1, we use the `offset` function. 

In our previous analysis of `Owls`, we modeled sibling negotiation on the other variables. Reading the paper, I believe that sibling negotiation is the total number of negotiations - not the number per chick. That means that we probably want to have an *offset*, which is the brood size. Let's run this model to see what we get. Remember, doing this means that we believe that the number of negotiations is proportional to the number of chicks. There are no synergies (where multiple chicks produce more negotiations than the sum of the chicks) or competitions (where multiple chicks produce fewer negotiations). Let's see how this model works.

```{r}
mod5 <- glmmTMB(
  sibling_negotiation ~ sex_parent + arrival_time + food_treatment + 
    offset(log(brood_size)) + (1 | nest),
  zif = ~ 1,
  family = nbinom2,
  data = oo
)
performance::check_zeroinflation(mod5)
summary(mod5)
```

Note that there is no estimate associated with `log(brood_size)`, because its coefficient is assumed to be one by the model. One way to test this (biological) assumption is to run the model with the log of brood size as a predictor, and see whether the coefficient is close to 1. 

```{r}
mod_test <- glmmTMB(
  sibling_negotiation ~ sex_parent + arrival_time + food_treatment +
    log(brood_size) + (1 | nest),
  zif = ~1,
  family = nbinom2,
  data = oo
)
summary(mod_test)
```

No, it is quite far from 1. This indicates that per-chick calling declines with larger broods.

