# Missing Data

In this shorter chapter, we discuss missing data. As a reminder, when there is missing data in a data frame, `lm` restricts to the complete cases. That is, `lm` runs a regression on only those observations for which there is no missing data in any of the variables. Depending on why the data is missing, this can lead to a loss of interpretive power, bias in the estimates of the coefficients, or worse outcomes. We will see examples of these (and simulations illustrating them) later in this chapter.

We start by discussing three types of missing data:

**Missing Completely at Random (MCAR)**. Data that is MCAR is randomly missing in a way that doesn't depend on any of observed or unobserved data. It is as if for each value in the data set, a random number was generated and the data point was removed if the random number satisfied a given property.

**Missing at Random (MAR)**. This is **unfortunately named** and is very often misunderstood. The probability that a value is missing **can** depend on the **other** data, but it doesn't depend on the value of the data itself. For example, blood pressure readings could be more often missing for younger patients, but once the probability of observing blood pressure is adjusted for age, the chance of it missing does not further depend on the blood pressure itself.

**Missing not at Random (MNAR)**. Data that is MNAR if the probability that it is missing depends on the values themselves, even after adjusting for the other observed variables. For example, if some patients with high blood pressure refuse to allow their blood pressure to be taken. If this refusal is not predictable via the other variables in the data set, then the missing blood pressure would be missing not at random.

There are two older techniques of dealing with missing data - complete case analysis (the default in `lm`) and mean imputation, where all missing values are replaced with the mean (or mode if categorical) of the data. Neither of these techniques are currently considered best practices for dealing with missing data.

Current best practices invlolve **imputing** the missing data on the observed variables **including the response**. Ideally, we should include the *uncertainty* in our imputation by repeating the imputation multiple times. We then run an analysis for each multiply imputed data set and combine the results. This technique is called **multiple imputation**. Finally, a **sensitivity analysis** is recommended, which will help explain how sensitive the model is to our missingness assumptions.

For those of you coming from a machine learning or predictive modeling background, including the response should cause alarm bells to ring. But, we will see that it is indeed a good idea for interpretive models. 

It is important to keep two things separate: **First** we will build an **imputation** model, which is used to impute missing data. **Second** we will build a model of the response, which is the interpretable  model that we are looking for in the end.

Let's see how all of this works with a specific data set. We consider the `Pima.tr2` data set from the `MASS` package.

```{r warning=FALSE, message=FALSE}
library(mice)
library(tidyverse)
pp <- MASS::Pima.tr2 |> 
  as_tibble()
pp
```

According to the documentation, a population of women who were at least 21 years old, of Pima Indian heritage and living near Phoenix, Arizona, was tested for diabetes according to World Health Organization criteria. The data were collected by the US National Institute of Diabetes and Digestive and Kidney Diseases.

```{r}
summary(pp)
```

We see that there are quite a few missing values. Suppose we wanted to determine whether the blood pressure depends on the diabetic criterion. Let's look at the pattern of missing data. 


```{r}
mice::md.pattern(pp)
```

We see that 84 observations are missing skin only, 1 is missing bp only, 12 are missing both skin and bp, etc. We certainly could just run a t-test on the 287 observed values. Realistically, that would be just fine. But, let's see how we can do multiple imputation in this context.


We start by building the multiple imputation model, using the default methods. Setting `m=20` ensures that we have 20 different data points imputed. A rule of thumb is that `m` should be as large as the percentage of observations with missing data, so in that case, we would set it equal to 33.

```{r}
imp <- mice(pp, m = 20, seed = 2026, printFlag = FALSE)
```

Next, we fit the model `bp ~ type` (which corresponds to a $t$-test but uses `lm`) using the imputed data.

```{r}
fit <- with(imp, lm(bp ~ type))
summary(pool(fit))
```

We see in the pooled model, we have a point of estimate of 4.22, so that women diabetis is associated with an increase of 4 units of blood pressure, and the $p$-value is $0.00371$. Let's compare that to what we would have obtained if we just used the complete cases.

```{r}
summary(lm(bp ~ type, data = pp))
```

Pretty similar. That is not surprising since there were only 13 missing values of bp. Note that in this case, we got a *smaller* effect size, a *larger* standard error, and a *higher* $p$-value. Imputing missing data does not always change your model in the direction of more significance; it is a principled way of trying to handle missing data.

## Technical Details

In this section, we discuss the technical details about how the imputation is made, and how the results are pooled to a single result. For simplicity, we will consider the following data set.

```{r}
dd <- data.frame(
  x1 = runif(40)
) |> 
  mutate(x2 = x1 + runif(40),
         y = 1 + 2 * x2 + rnorm(40))

dd$x1[1:3] <- NA
dd$x2[c(1, 5)] <- NA
```

We have 5 missing values, distributed as follows:

```{r}
md.pattern(dd)
```

We start by assigning values to the missing values of `x1` and `x2`.

```{r}
aa <- dd |> 
  mutate(x1 = ifelse(is.na(x1), sample(x1[!is.na(x1)], size = 3, replace = T), x1),
         x2 = ifelse(is.na(x2), sample(x1[!is.na(x2)], size = 3, replace = T), x2))
md.pattern(aa)
```

It is unlikely that these are good choices for `x1` and `x2`, but since they are sampled from existing values of the variables, at least we know they are **possible** values. For example, we know (because of the construction) that `x1` needs to be between 0 and 1, and it will be for these poorly imputed values.

Next, we model `x1` on the other variables.

```{r}
missingx1 <- 1:3
missingx2 <- c(1, 5)
mod <- lm(x1 ~ y + x2, data = aa)
```

Now, we need to choose a value of `x1` that is close to the predicted value with some noise added in. We are not going to follow the exact procedure of `mice` here, but simplify it as follows. The way we add noise is by adding it to the **coefficients** rather than the predicted values. In this case, we will take a random draw from a multivariate normal distribution with mean of the predicted values of the coefficients and covariance matrix given by our estimate of that covariance matrix from `lm`. 

```{r}
betadot <- MASS::mvrnorm(1, mu = coef(mod), Sigma = vcov(mod))
betadot
coef(mod)
```

Now we predict our `x1` variable from the perturbed values of the coefficients. We can compare to the prediction from the original value of the coefficients by using `predict`. 

```{r}
pp <- betadot %*% t(matrix(c(rep(1, times = length(missingx1)), aa$y[missingx1], aa$x2[missingx1]), nrow = length(missingx1)))
pp
predict(mod, newdata = aa[missingx1,])
```

We don't automatically take the closest value to the predicted values; we choose the nearest 5 or so closest values and choose one at random.

```{r}
find_replacement <- function(vector, val) {
  candidates <- sort(abs(vector - val), index.return = TRUE)$ix[1:5]
  vector[sample(candidates, 1)]
}

find_replacement <- Vectorize(find_replacement, vectorize.args = "val")
aa$x1[missingx1] <- find_replacement(aa$x1, pp)
```

Now, we repeat the entire process for the second predictor variable.

```{r}
mod <- lm(x2 ~ y + x1, data = aa)
betadot <- MASS::mvrnorm(1, mu = coef(mod), Sigma = vcov(mod))
pp <- betadot %*% t(matrix(c(rep(1, times = length(missingx1)), aa$y[missingx1], aa$x2[missingx1]), nrow = length(missingx1)))
find_replacement <- Vectorize(find_replacement, vectorize.args = "val")
aa$x2[missingx1] <- find_replacement(aa$x2, pp)
```

Then, we would repeat this a bunch of times for a burn-in or initialization phase, and after that burn-in, we would start to take values. The package name `mice` stands for multiple imputation by chained equations; the chaining is the repeating of the imputation like we are doing here. 

To be clear, this leaves out some important steps along the way, but it does give the **flavor** of the algorithm.  

Now that we have seen how it works for **one stream**, we need to get multiple imputations. `mice` will have as many streams as you want imputations, and after a number of iterations, it will take one data set from each stream. For diagnostics, what we like to see is that each of the streams is mixing well. They are offering similar ranges of values across each of the streams. You don't want to see one stream off in Jefferson City while the other streams are in CWE. In practice, we will use the built in diagnostics to determine whether that is accurate.

## Diagnostics

We return to our example above and look at the built in diagnostics. We re-run the imputation model to include more iterations and more streams.

```{r}
pp <- MASS::Pima.tr2 |> 
  as_tibble()
imp <- mice(pp, m = 10, maxit = 100, seed = 2026, printFlag = FALSE)
plot(imp)
```

We can see that the chains seem to be mixing pretty well over the ten iterations shown. Now, we check convergence.

```{r}
diag_mean <- convergence(imp, diagnostic = "all", parameter = "mean")
diag_mean |> 
  filter(!is.na(ac)) |> 
  head()

diag_sd <- convergence(imp, diagnostic = "all", parameter = "sd")
diag_sd |> 
  filter(!is.na(ac)) |> 
  head()
```

The `psrf` variable is measuring the variance of the variable *within* the stream versus the variance of the variable *between* the streams. Ideally, this should be close to 1. The `ac` variable is measuring the autocorrelation of the variable. Ideally, this should be close to 0. However, possibly more important than the *values* of the statistics is the *behavior* of the statistics. If the statistic shows a clear decreasing trend over iterations, then that is evidence that the stream has not yet fully converged, and we should consider running the stream for more iterations. Let's look at some plots. 

```{r}
# Plot lag-1 autocorrelation for chain means
diag_mean
diag_mean |> 
  filter(str_detect(vrb, "bp|skin|bmi")) |> 
  ggplot(aes(x = .it, y = ac)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~ vrb, scales = "free_y") +
  labs(
    x = "Iteration",
    y = "Lag-1 autocorrelation",
    title = "MICE convergence diagnostics: chain means"
  )
```

For the first plot, we see a decreasing lag-1 autocorrelation through about the first 60 iterations, at which point it starts to wander. We see that `skin` and `bmi` neither get to 0 autocorrelation, but they seem to have stabilized. We present the rest of the plots without comment.

```{r}
# Plot PSRF for chain means
diag_mean |> 
  filter(str_detect(vrb, "bp|skin|bmi")) |> 
  ggplot(aes(x = .it, y = psrf)) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = 2) +
  facet_wrap(~ vrb, scales = "free_y") +
  labs(
    x = "Iteration",
    y = "PSRF",
    title = "MICE convergence diagnostics: chain means"
  )

# Repeat for chain SDs if you want
diag_sd |> 
  filter(str_detect(vrb, "bp|skin|bmi")) |> 
  ggplot(aes(x = .it, y = ac)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~ vrb, scales = "free_y") +
  labs(
    x = "Iteration",
    y = "Lag-1 autocorrelation",
    title = "MICE convergence diagnostics: chain SDs"
  )

diag_sd |> 
  filter(str_detect(vrb, "bp|skin|bmi")) |> 
  ggplot(aes(x = .it, y = psrf)) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = 2) +
  facet_wrap(~ vrb, scales = "free_y") +
  labs(
    x = "Iteration",
    y = "PSRF",
    title = "MICE convergence diagnostics: chain SDs"
  )
```

Overall, the convergence looks good. We do have some autocorrelation in the successive estimates of variables. According to classical MCMC diagnostics, autocorrelation generally impacts the **effective sample size**. It is not clear whether that is an issure here, though, as we aren't taking multiple imputations from the same stream.

Overall, these diagnostics look pretty good, and if we like, we can run the full analysis twice to see whether we get similar results as a final test.

```{r}
imp2 <- mice(pp, m = 10, maxit = 100, set.seed = 2027, printFlag = FALSE)
fit <- with(imp, lm(bp ~ type))
summary(pool(fit))
fit2 <- with(imp2, lm(bp ~ type))
summary(pool(fit2))
```

If we want to see confidence intervals after pooling, we use:

```{r}
summary(pool(fit), conf.int = T)
```


## Pooling

Now that we understand how the multiple imputations are produced, we need to understand how the results are pooled into a final answer. We only consider OLS. The coefficient values are easy: you estimate the coefficient for each imputed data set and take the mean.

The first command creates a **list** of all of the data sets. We can run each one of those through `lm`.

```{r}
datlist <- complete(imp, action = "all")

fits <- purrr::map_df(datlist, function(dat) {
  lm(bp ~ type, data =dat) |> 
    broom::tidy()
})

head(fits)
```

We see we have slope and intercept estimates for each of the imputed data sets. To get the pooled estimate, we simply average and compare to `mice::pool`.

```{r}
fits |> 
  group_by(term) |> 
  summarise(estimate = mean(estimate)) |> 
  as.data.frame() |> 
  print(digits = 7)

pool(with(imp, lm(bp ~ type))) |> summary()
```

For the pooled estimate of standard error for the slope, we need to combine both an estimate of the the standard errors and additionally we need to add a component that takes into account our uncertainty in the estimate of the coefficients. I asked ChatGPT to give a short heuristic for the formula, and it wrote - Rubin’s pooling rule adds the between-imputation variance \(B\) because the estimates from the different imputed datasets do not agree exactly, and that disagreement reflects real uncertainty about the missing values. The usual within-imputation variance \(\bar U\) tells us how uncertain we would be if the data were complete, while \(B\) measures the additional uncertainty created by not knowing the missing data. Thus the pooled variance must account for both ordinary sampling variability and the extra variability introduced by the imputation process.


```{r}
pool(with(imp, lm(bp ~ type))) 
fits |> 
  filter(term == "typeYes") |> 
  summarize(B = var(estimate), Ubar = mean(std.error^2)) |> 
  summarize(se = sqrt((1 + 1/10) * B + Ubar)) |> 
  as.data.frame() |> 
  print(digits = 7)

pool(with(imp, lm(bp ~ type))) |> summary()
```

## Simulations

Now, let's compare using `mice` to using complete cases and using mean imputation, two methods that are also relatively commonly used. We compare the bias of estimates 95 for the coefficients in the three cases. 

We will have a data set with two predictors and a response, and we will choose missing data according to other data in the set. Note that this is essentially the model that we use in `mice`. If the data is missing by a different mechanism, then results might be different.

```{r}
dd <- data.frame(x1 = runif(50),
           x2 = runif(50))

dd <- dd |> 
  mutate(y = x1 + 2 * x2 + rnorm(50, 0, 1.5)) 
```

That is our data set, now we need to inject missing data. We will aim for about 10% missing in `x1` and `x2` separately. Some rows may have both variables missing.

```{r}

rep <- TRUE
while(rep) {
  b1 <- runif(1, -5, 0)
  b2 <- runif(1, -20, 0)
  ex <- sum(plogis(b1 * dd$y + b2 * dd$x2))
  if(ex > 4 && ex < 6) {
    rep <- FALSE
  }
}
removex1 <- which(as.logical(rbinom(50, 1, prob = plogis(b1 * dd$y + b2 * dd$x2))))


rep <- TRUE
while(rep) {
  b1 <- runif(1, -5, 0)
  b2 <- runif(1, -10, 0)
  ex <- sum(plogis(b1 * dd$y + b2 * (dd$x1 + 0.5)))
  plogis(b1 * dd$y + b2 * (dd$x1 + 0.5))
  ex
  if(ex > 4 && ex < 6) {
    rep <- FALSE
  }
}
removex2 <- which(as.logical(rbinom(50, 1, prob = plogis(b1 * dd$y + b2 * (dd$x1 + 0.5)  ))))

dd$x1[removex1] <- NA
dd$x2[removex2] <- NA
```

Our three models are built below:

```{r}
mod1 <- lm(y ~ x1 + x2, data = dd)
broom::tidy(mod1) |> 
  select(term, estimate) |> 
  mutate(method = "complete_cases")

dd2 <- dd |> 
  mutate(x1 = case_when(
    is.na(x1) ~ mean(x1, na.rm = T),
    TRUE ~ x1
  ),
  x2 = case_when(
    is.na(x2) ~ mean(x2, na.rm = T),
    TRUE ~ x2
  ))
mod2 <- lm(y ~ x1 + x2, data = dd2)
broom::tidy(mod2) |> 
  select(term, estimate) |> 
  mutate(method = "mean")

imp <- mice(dd, m = 12, printFlag = F)
mod3 <- with(imp, lm(y ~ x1 + x2))
pool(mod3)$pooled |> 
  select(term, estimate) |> 
  mutate(method = "mice")

```

Now we iterate:

```{r eval=FALSE}
tt <- purrr::map_df(1:500, function(x) {
  dd <- data.frame(x1 = runif(50),
                   x2 = runif(50))
  
  dd <- dd |> 
    mutate(y = x1 + 2 * x2 + rnorm(50, 0, 1.5)) 
  
  rep <- TRUE
  while(rep) {
    b1 <- runif(1, -5, 0)
    b2 <- runif(1, -20, 0)
    ex <- sum(plogis(b1 * dd$y + b2 * dd$x2))
    if(ex > 3 && ex < 7) {
      rep <- FALSE
    }
  }
  removex1 <- which(as.logical(rbinom(50, 1, prob = plogis(b1 * dd$y + b2 * dd$x2))))
  
  
  rep <- TRUE
  while(rep) {
    b1 <- runif(1, -5, 0)
    b2 <- runif(1, -10, 0)
    ex <- sum(plogis(b1 * dd$y + b2 * (dd$x1 + 0.5)))
    plogis(b1 * dd$y + b2 * (dd$x1 + 0.5))
    ex
    if(ex > 3 && ex < 7) {
      rep <- FALSE
    }
  }
  removex2 <- which(as.logical(rbinom(50, 1, prob = plogis(b1 * dd$y + b2 * (dd$x1 + 0.5)  ))))
  
  dd$x1[removex1] <- NA
  dd$x2[removex2] <- NA
  
  
  mod1 <- lm(y ~ x1 + x2, data = dd)
  
  
  dd2 <- dd |> 
    mutate(x1 = case_when(
      is.na(x1) ~ mean(x1, na.rm = T),
      TRUE ~ x1
    ),
    x2 = case_when(
      is.na(x2) ~ mean(x2, na.rm = T),
      TRUE ~ x2
    ))
  mod2 <- lm(y ~ x1 + x2, data = dd2)
  
  
  imp <- mice(dd, m = 12, printFlag = F)
  mod3 <- with(imp, lm(y ~ x1 + x2))
  
  bind_rows(broom::tidy(mod2) |> 
              select(term, estimate) |> 
              mutate(method = "mean"),
            broom::tidy(mod1) |> 
              select(term, estimate) |> 
              mutate(method = "complete_cases"),
            pool(mod3)$pooled |> 
              select(term, estimate) |> 
              mutate(method = "mice")) |> 
    mutate(iteration = x)
  
  
})

save(tt, file = "mice_simulation.rda")
```

```{r}
load("mice_simulation.rda")

tt |> 
  group_by(term, method) |> 
  summarize(mu = mean(estimate)) |> 
  arrange(method)
```

## Sensitivity Analysis

Let's return to the Pima blood pressure data set. Multiple imputation has an assumption about the missing data mechanism, and in our case that assumption is MAR. It is a good idea to ask how much our final conclusion would change if the missing blood pressure values tended to be systematically higher or lower than the MAR imputations suggest. For example, perhaps people with higher blood pressure than would be predicted by the other variables are more likely to refuse a blood pressure reading.

One simple way to do this is with a \(\delta\)-adjustment. We let `mice` impute `bp` as usual, and then during each iteration we add a constant \(\delta\) to the imputed blood pressure values. The case \(\delta = 0\) is the original MAR analysis. Negative values of \(\delta\) force the missing blood pressure values to be lower than the MAR imputations, while positive values force them to be higher.

We return to the `Pima.tr2` data and consider a range of values of \(\delta\).

```{r}
pp <- MASS::Pima.tr2 |> 
  as_tibble()
```

The method for adjustment after each iteration is quite general, and is handled via a `post` argument to the `mice` function. Here is the syntax of what we will pass to `mice`:

```{r}
delta <- 5
post <- make.post(pp)
post["bp"] <- paste0("imp[[j]][, i] <- imp[[j]][, i] + ", delta) 
post
```

This somewhat weird notation indicates that for each imputation chain, we add 5 to the ith observation after each time that it is imputed. Note that we do not wait until the imputation chain terminates to add 5, because the additional 5 could propogate through the other variables and make things vary more (or less) after propogation.

We will not hard code `delta = 5` but write a function (with help from our friends at chat) to do this for general `delta`.

```{r}
run_delta <- function(delta) {
  post <- make.post(pp)
  post["bp"] <- paste0("imp[[j]][, i] <- imp[[j]][, i] + ", delta) 
  
  imp <- mice(pp, m = 20, maxit = 20, seed = 2026, 
              post = post, printFlag = FALSE)
  
  with(imp, lm(bp ~ type)) |> 
    pool() |> 
    summary() |> 
    filter(term == "typeYes") |> 
    mutate(delta = delta,
           conf.low = estimate - qt(0.975, df) * std.error,
           conf.high = estimate + qt(0.975, df) * std.error) |> 
    select(delta, estimate, std.error, conf.low, conf.high, p.value)
}

deltas <- c(-8, -4, 0, 4, 8)

sens <- purrr::map_df(deltas, run_delta)

sens |> 
  as.data.frame() |> 
  print(digits = 7)
```

The row with `delta = 0` is the original MAR analysis. The other rows tell us what happens if the missing blood pressure values are systematically lower or higher than the MAR model suggests. Note that in this case, the primary conclusion (that diabetes is associated with higher blood pressure) is not changed. It does change the point estimate of the effect as well as the confidence intervals. One might write that a $\delta$ sensitivity analysis was performed and the effect size exhibited moderate sensitivity to departure from MAR, ranging from 3.89 to 4.69. The direction and significance of the effect remained the same throughout the range of $\delta$ that were tested.

A plot is sometimes easier to interpret than a table.

```{r}
sens |> 
  ggplot(aes(x = delta, y = estimate)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.4) +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(
    x = expression(delta),
    y = "Estimated coefficient for typeYes",
    title = "Sensitivity analysis for departures from MAR"
  )
```

There is no single correct choice of \(\delta\); it should be guided by subject matter knowledge. The point of the analysis is to see whether our conclusion is stable over a range of plausible departures from MAR. If the estimated coefficient for `typeYes` stays roughly the same over a reasonable range of values of \(\delta\), then our conclusion is not especially sensitive to the missingness assumption. If it changes a great deal, or even changes sign, then the conclusion is more fragile than the original MAR analysis alone would suggest.

This is not the only way to do a sensitivity analysis, but it is a simple and useful first approach.





