# Regression Adjustment and Causal Interpretation

In this chapter, we begin our discussion of causal inference. We will not try to cover the subject in full here. Instead, we focus on a basic but important question: when we fit a regression model with a treatment variable, under what circumstances can we interpret the coefficient of that treatment variable causally?

A useful place to start is with observational data. In observational studies, the treated and untreated groups are often different before the treatment occurs. This means that a simple comparison of outcomes can mix together two things: the effect of the treatment itself, and baseline differences between the groups. Regression adjustment is one way to address this problem, but it does not solve it automatically.

We will work with the `lalonde` data set from the `MatchIt` package. The outcome variable is earnings in 1978 (`re78`), and the treatment variable `treat` indicates participation in a job-training program. The data also includes several pre-treatment covariates such as age, education, marital status, degree status, and earnings in 1974 and 1975.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(MatchIt)
library(broom)

aa <- MatchIt::lalonde |> 
  as_tibble()
```

We begin by looking at the structure of the data.

```{r}
aa |> 
  select(treat, age, educ, married, nodegree, re74, re75, re78) |> 
  summary()
```

We start with the simplest possible regression model,

\[
\texttt{re78} = \beta_0 + \beta_1 \texttt{treat} + \varepsilon.
\]

This model compares 1978 earnings between the treated and untreated groups without adjusting for any other variables.

```{r}
mod_naive <- lm(re78 ~ treat, data = aa)

mod_naive |>
  tidy(conf.int = TRUE)
```

It is also useful to look at the group means directly.

```{r}
aa |> 
  group_by(treat) |> 
  summarise(mean_re78 = mean(re78), .groups = "drop")
```

Note that getting the job training is not statistically significant, and the point estimate of the effect of training is **negative** 635 dollars. That means that the people who received traning had salaries on average *lower* than those who did not. One might conclude that the program did not help. However, this conclusion would be too quick. The data are observational, not randomized, so it is entirely possible that the treated and untreated groups differed before the program began.

### Baseline Differences Between the Groups

To investigate this, we compare some pre-treatment covariates across the two groups.

```{r}
aa |> 
  group_by(treat) |> 
  summarise(
    across(c(age, educ, re74, re75), mean),
    married = mean(married),
    nodegree = mean(nodegree),
    .groups = "drop"
  )
```

We can also visualize one of the most important pre-treatment variables, namely earnings in 1975.

```{r}
aa |> 
  mutate(treat = factor(treat, levels = c(0, 1), labels = c("Control", "Treated"))) |> 
  ggplot(aes(x = treat, y = re75)) +
  geom_boxplot() +
  labs(x = NULL, y = "1975 earnings")
```

We see that the treated and untreated groups are not the same at baseline. In particular, prior earnings differ substantially. This matters because prior earnings are measured before treatment, and they are strongly related to later earnings. If the treated group started out at a disadvantage, then a raw treated-versus-control comparison is not isolating the effect of treatment. It is combining the effect of treatment with the fact that the treated and untreated groups were different to begin with.

This is the basic idea of **confounding**. A confounder is a variable that is related both to treatment assignment and to the outcome. Prior earnings are a natural example in this setting.

A natural first thing to do is to control for the observed pre-treatment covariates in a regression model. For example, we can fit

\[
\texttt{re78} = \beta_0 + \beta_1 \texttt{treat} + \beta_2 X + \varepsilon,
\]

where $X$ denotes the baseline covariates.

```{r}
mod_adj <- lm(
  re78 ~ treat + age + educ + race + married + nodegree + re74 + re75,
  data = aa
)

mod_adj |>
  tidy(conf.int = TRUE)
```

The treatment coefficient changes noticeably after adjustment. Now, it is statistically significant and **positive**. Our estimate is that the job training program is associated with an increase of 1548 dollars after adjusting for the other covariates.

Regression adjustment is therefore useful. It can remove bias due to **observed** baseline differences, at least when the model is doing a reasonable job of adjustment. At the same time, it is important not to say more than the model justifies. It would **not** be justified to say that participation in the pogram **increased** 1978 earnings, because that is assigning causality to the program. All we know is that it is **associated** with an increase after adjusting for the other covariates. In order to justify a causality claim, we would need to believe that after controlling for the observed pre-treatment covariates, the treated and untreated groups are comparable in all ways relevant to the outcome.

This is a much stronger claim than simply saying that we included several variables in a regression. A regression coefficient is not automatically causal just because it comes from a multiple regression model. The causal interpretation comes from the assumptions we are willing to defend.

A good way to summarize the distinction is the following:

- The naive model compares treated and untreated units directly.
- The adjusted model compares treated and untreated units after controlling for observed baseline covariates.
- A causal interpretation requires that this adjusted comparison be a good stand-in for the missing *counterfactual* comparison, see below for details on counterfactuals.

One way to formalize the causal question is through the potential outcomes framework. For each individual $i$, let

- $Y_i(1)$ be the earnings that person $i$ would have if treated;
- $Y_i(0)$ be the earnings that person $i$ would have if untreated.

The individual causal effect is then

\[
Y_i(1) - Y_i(0).
\]

The difficulty is that we never observe both values for the same person. If person $i$ is treated, then we observe $Y_i(1)$ but not $Y_i(0)$. If person $i$ is untreated, then we observe $Y_i(0)$ but not $Y_i(1)$. The observed outcome can be written as

\[
Y_i = D_iY_i(1) + (1-D_i)Y_i(0),
\]

where $D_i$ is the treatment indicator. That is, $D_i = 1$ if treatment is received and $D_i = 0$ if no treatment is received.

This is the fundamental problem of causal inference. For each person, one of the two relevant outcomes is missing. In observational data, we use untreated individuals as stand-ins for the missing untreated outcomes of the treated group, or treated individuals as stand-ins for the missing treated outcomes of the untreated group. The central question is whether those stand-ins are credible.

In the present example, the raw comparison is not especially credible because the groups differ so much at baseline. The adjusted regression is more credible, but only to the extent that the included covariates capture the important differences between the two groups. In particular, there is nothing about the multiple regression that directly tries to make the two groups comparable; if it happens, it is fortuitous. 

Later, we will study a different way to organize this comparability problem. Rather than putting all of the adjustment into a single outcome regression, we can first ask how likely each unit was to receive treatment based on the observed pre-treatment covariates.

This leads to the **propensity score**, defined by

\[
e(X) = P(D = 1 \mid X).
\]

Here, $X$ denotes the observed pre-treatment covariates. Units with similar propensity scores have similar observed baseline characteristics, at least in the sense relevant for treatment assignment. The idea is then to compare treated and untreated units that were similarly likely to receive treatment in the first place.

We will not develop propensity scores fully in this section, but we can at least estimate them and inspect the results. A common first step is to fit a logistic regression for treatment.

```{r}
ps_mod <- glm(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = aa,
  family = binomial()
)

aa_ps <- ps_mod |>
  augment(data = aa, type.predict = "response") |>
  mutate(
    treat = factor(treat, levels = c(0, 1), labels = c("Control", "Treated"))
  )

ps_mod |>
  tidy()
```

We can now look at the estimated propensity scores by treatment group.

```{r}
aa_ps |> 
  ggplot(aes(x = .fitted, fill = treat)) +
  geom_histogram(alpha = 0.5, bins = 30, position = "identity") +
  labs(x = "Estimated propensity score", y = "Count", fill = NULL)
```

The plot is not itself a causal analysis, but it begins to show us something that the outcome regression did not show directly: how much overlap there is between the treated and untreated groups on the observed baseline covariates. Later, this idea of overlap will become central. In this particular plot, notice that there is not a lot of overlap between control and treated in the low propensity score group. This means that there is some combination of covariates which almost uniformly led participants not to get job training. That is somewhat bad news for causal analysis in general, and certainly should lead to hesitancy in interpreting the multiple regression output causally. We expect to see some predictive power of the covariates on whether the treatment was received, but we are hoping to still see significant overlap for our causal analysis. There are techniques for dealing with data like this which we will discuss in later sections.

### Summary

This section makes three points.

First, a simple regression of outcome on treatment can be misleading in observational data because treated and untreated groups often differ before treatment occurs.

Second, regression adjustment can account for observed baseline differences and can change the treatment estimate substantially. In the job-training example, this is exactly what happens.

Third, even an adjusted regression is not automatically causal. The causal interpretation requires additional assumptions about comparability between the treated and untreated groups after adjustment.

Propensity scores give us another way to think about that comparability. They do not remove the need for assumptions, but they make the adjustment problem more explicit. In the next part of our causal inference discussion, we will use them to study matching and related methods.

## Identification Assumptions for Causal Adjustment

In the previous section, we saw that a naive regression of later earnings on treatment in the Lalonde data gave a different answer than a regression that adjusted for observed baseline covariates. That already tells us something important: the treated and control groups were different before treatment. In observational work, that is the rule rather than the exception.

The next question is the more important one. When does an adjusted comparison deserve a causal interpretation? Regression adjustment can be part of a causal analysis, but the causal interpretation does not come from the software output. It comes from the assumptions that justify using treated and untreated units as stand-ins for one another's missing counterfactual outcomes.

Throughout this section, we continue to use the Lalonde job-training data. The treatment is participation in the job-training program, and the outcome is earnings in 1978.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(MatchIt)
library(broom)

data("lalonde")
lalonde

aa <- lalonde |>
  as_tibble() |>
  mutate(
    treat = factor(treat, levels = c(0, 1), labels = c("Control", "Treated")),
    race = factor(race),
    married = factor(married, levels = c(0, 1), labels = c("No", "Yes")),
    nodegree = factor(nodegree, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

We will use the following observed pre-treatment covariates throughout this section:

- `age`
- `educ`
- `race``
- `married`
- `nodegree`
- `re74`
- `re75`

### Assumptions for Causal Adjustment

We continue to write the potential outcomes for person $i$ as $Y_i(1)$ and $Y_i(0)$, where these represent that person's 1978 earnings under treatment and under control, respectively. The observed outcome is

$$
Y_i = D_iY_i(1) + (1-D_i)Y_i(0).
$$

This is the formal statement of **consistency**. If a person is treated, then the observed outcome is the treated potential outcome; if a person is untreated, then the observed outcome is the untreated potential outcome. This sounds automatic, but it depends on the treatment being defined clearly enough that $Y(1)$ and $Y(0)$ are meaningful. If the treatment were simply coded as "exercise," then the treated condition might bundle together walking, weight lifting, marathon training, yoga, and many other interventions. In that situation, the meaning of $Y(1)$ becomes murky. Likewise, if "job training" bundled together many radically different programs, then the treated potential outcome would be less clearly defined.

The most important assumption in observational work is **ignorability**:

$$
(Y(1), Y(0)) \perp D \mid X.
$$

In words, after conditioning on the observed pre-treatment covariates $X$, treatment assignment is independent of the potential outcomes. A slightly more informal version is often easier to remember: once we control for the relevant observed pre-treatment variables, treated and untreated units are comparable on average. This assumption is also called **no unmeasured confounding** or **selection on observables**.

It is important to be clear about what ignorability does and does not say. It does not say that treatment is random in the raw data. It says that treatment is *as if random after conditioning on the chosen covariates*. That is why the choice of covariates matters so much.

This connects directly to the informal **backdoor idea**. If a variable is a common cause of treatment and outcome, then failing to adjust for it creates a noncausal association between treatment and outcome. In the Lalonde data, prior earnings are a natural example. Someone's 1975 earnings may affect whether they enter the training program, and 1975 earnings will also help predict 1978 earnings. In that case, `re75` is a confounder, and adjusting for it blocks a noncausal path between treatment and outcome.

A third important assumption is **positivity**, sometimes called **overlap**:

$$
0 < P(D = 1 \mid X = x) < 1
$$

for the values of $x$ we care about. In plain language, this means that for covariate profiles of interest, there must be some chance of both treatment and control. If units with certain covariates are always treated, or always untreated, then we do not have comparable treated and control units in that part of the covariate space. In practice, this means that causal adjustment requires overlap. No overlap means no apples-to-apples comparison.

A final assumption that is worth stating is **SUTVA**, but in this chapter we will emphasize only the part that is genuinely new after consistency: **no interference**. One person's treatment should not directly affect another person's outcome. In the job-training setting, this could fail if treated and untreated workers are competing for the same jobs in the same market. We will not focus on interference in this chapter, but it is worth keeping in mind.

These assumptions help explain why regression is not the opposite of causal inference. Regression is one tool that causal inference can use. The causal interpretation depends on whether the regression is implementing a defensible comparison.

There is one more practical point to make before returning to the Lalonde data. A good control is generally a variable measured before treatment that affects both treatment assignment and the outcome. A bad control is a variable that is downstream of treatment or a collider. For example, if a job-search intervention improves later mental health partly by increasing job-search confidence, then controlling for job-search confidence would remove part of the very treatment effect we are trying to estimate. If the goal is the total effect of treatment, post-treatment mediators are bad controls.

The `jobs` data in the `mediation` package give a nice example. The treatment variable is `treat`, the outcome is `depress2`, and `job_seek` is a plausible post-treatment mediator. In the code below, the third model adds `job_seek` after treatment has already occurred.

```{r}
library(mediation)

data("jobs")

jobs2 <- jobs |>
  as_tibble()

mod1 <- lm(depress2 ~ treat, data = jobs2)
mod2 <- lm(depress2 ~ treat + econ_hard + sex + age, data = jobs2)
mod3 <- lm(depress2 ~ treat + econ_hard + sex + age + job_seek, data = jobs2)

bind_rows(
  tidy(mod1) |> mutate(model = "Treatment only"),
  tidy(mod2) |> mutate(model = "Baseline controls"),
  tidy(mod3) |> mutate(model = "Adds post-treatment variable")
) |> 
  filter(term == "treat") |> 
  dplyr::select(model, estimate, std.error, statistic, p.value)
```

When the coefficient on treatment changes after `job_seek` is added, that is not automatically an improvement. If `job_seek` lies on the causal pathway from treatment to the outcome, then controlling for it changes the estimand. This is why "control for everything" is not safe advice.

### The Lalonde Data Revisited

We return now to the Lalonde data. The first thing to notice is that the treated and control groups differ on several observed pre-treatment covariates.

```{r}
aa |>
  group_by(treat) |>
  summarize(
    n = n(),
    age = mean(age),
    educ = mean(educ),
    re74 = mean(re74),
    re75 = mean(re75),
    .groups = "drop"
  )
```

```{r}
aa |>
  group_by(treat) |>
  summarize(
    percent_black = mean(race == "black"),
    percent_hispanic = mean(race == "hispan"),
    percent_white = mean(race == "white"),
    married = mean(married == "Yes"),
    nodegree = mean(nodegree == "Yes"),
    .groups = "drop"
  )
```

These are all pre-treatment variables, so differences here are a warning sign for confounding rather than an effect of treatment. In particular, prior earnings differ between the groups, and prior earnings are highly relevant for later earnings.

We can see again that the naive and adjusted regressions tell different stories.

```{r}
mod_naive <- lm(re78 ~ treat, data = aa)

mod_adj <- lm(
  re78 ~ treat + age + educ + race + married + nodegree + re74 + re75,
  data = aa
)

bind_rows(
  tidy(mod_naive, conf.int = TRUE) |> mutate(model = "Naive"),
  tidy(mod_adj, conf.int = TRUE) |> mutate(model = "Adjusted")
) |>
  filter(term == "treatTreated") |>
  dplyr::select(model, estimate, conf.low, conf.high, p.value)
```

The sign change is a strong indication that the raw treatment comparison was confounded by observed baseline differences. This does not prove that the adjusted model is causal, but it does show that adjustment matters.

It is useful to visualize overlap directly. We start with one important pre-treatment covariate, 1975 earnings.

```{r fig.height=4}
ggplot(aa, aes(x = re75, fill = treat)) +
  geom_density(alpha = 0.4) +
  labs(x = "1975 earnings", y = "Density", fill = NULL)
```

Where the two distributions overlap, comparison is more plausible. Where they are far apart, adjustment begins to rely on extrapolation. This same idea can be summarized more compactly by the **propensity score**, which we define as

$$
e(X) = P(D = 1 \mid X).
$$

For now, we use the propensity score only as a way to study overlap. We estimate it with logistic regression using the observed pre-treatment covariates.

```{r}
ps_mod <- glm(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = aa,
  family = binomial()
)

aa_ps <- ps_mod |>
  augment(data = aa, type.predict = "response") |>
  mutate(ps = .fitted) |>
  dplyr::select(-.fitted)
```

Now we look at the distributions of the estimated propensity scores in the treated and control groups.

```{r fig.height=4}
ggplot(aa_ps, aes(x = ps, fill = treat)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 20) +
  labs(x = "Estimated propensity score", y = "Count", fill = NULL)
```

The treated units tend to have higher estimated probabilities of treatment than the control units. This is the expected direction. The issue is not that the graph is backwards. The issue is that the groups may be too separated. Good causal adjustment needs overlap, not simply a treatment model that predicts treatment well. If treated and control units have little overlap in propensity score, then there are few comparable units to compare.

This is the main reason to introduce propensity scores in the first place. Regression adjustment hides the comparability problem inside a single coefficient. Propensity scores make it easier to inspect whether treated and untreated units are similar on the observed pre-treatment covariates and whether common support is adequate.

The big lesson from this section is that adjusted comparisons only deserve a causal interpretation under assumptions. Consistency tells us that the treatment must be defined clearly enough for the potential outcomes to make sense. Ignorability says that after conditioning on the observed pre-treatment covariates, treatment assignment is as good as random. Positivity says that comparable treated and control units must exist. Finally, the discussion of bad controls reminds us that adjustment can go wrong when we condition on variables that are affected by treatment. The next section uses propensity scores more directly to organize causal comparisons in observational data.
