# Causal Inference

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

In this chapter, we begin our discussion of causal inference. We will not try to cover the subject in full here. Instead, we focus on a basic but important question: when we fit a regression model with a treatment variable, under what circumstances can we interpret the coefficient of that treatment variable causally?

A useful place to start is with observational data. In observational studies, the treated and untreated groups are often different before the treatment occurs. This means that a simple comparison of outcomes can mix together two things: the effect of the treatment itself, and baseline differences between the groups. Regression adjustment is one way to address this problem, but it does not solve it automatically.

We will work with the `lalonde` data set from the `MatchIt` package. The outcome variable is earnings in 1978 (`re78`), and the treatment variable `treat` indicates participation in a job-training program. The data also includes several pre-treatment covariates such as age, education, marital status, degree status, and earnings in 1974 and 1975.

```{r warning=FALSE, message=FALSE}
library(MatchIt)
library(tidyverse)
library(broom)



aa <- MatchIt::lalonde |> 
  as_tibble()
```

We begin by looking at the structure of the data.

```{r}
aa |> 
  select(treat, age, educ, married, nodegree, re74, re75, re78) |> 
  summary()
```

We start with the simplest possible regression model,

\[
\texttt{re78} = \beta_0 + \beta_1 \texttt{treat} + \varepsilon.
\]

This model compares 1978 earnings between the treated and untreated groups without adjusting for any other variables.

```{r}
mod_naive <- lm(re78 ~ treat, data = aa)

mod_naive |>
  tidy(conf.int = TRUE)
```

It is also useful to look at the group means directly.

```{r}
aa |> 
  group_by(treat) |> 
  summarise(mean_re78 = mean(re78), .groups = "drop")
```

Note that getting the job training is not statistically significant, and the point estimate of the effect of training is **negative** 635 dollars. That means that the people who received traning had salaries on average *lower* than those who did not. One might conclude that the program did not help. However, this conclusion would be too quick. The data are observational, not randomized, so it is entirely possible that the treated and untreated groups differed before the program began.

### Baseline Differences Between the Groups

To investigate this, we compare some pre-treatment covariates across the two groups.

```{r}
aa |> 
  group_by(treat) |> 
  summarise(
    across(c(age, educ, re74, re75), mean),
    married = mean(married),
    nodegree = mean(nodegree),
    .groups = "drop"
  )
```

We can also visualize one of the most important pre-treatment variables, namely earnings in 1975.

```{r}
aa |> 
  mutate(treat = factor(treat, levels = c(0, 1), labels = c("Control", "Treated"))) |> 
  ggplot(aes(x = treat, y = re75)) +
  geom_boxplot() +
  labs(x = NULL, y = "1975 earnings")
```

We see that the treated and untreated groups are not the same at baseline. In particular, prior earnings differ substantially. This matters because prior earnings are measured before treatment, and they are strongly related to later earnings. If the treated group started out at a disadvantage, then a raw treated-versus-control comparison is not isolating the effect of treatment. It is combining the effect of treatment with the fact that the treated and untreated groups were different to begin with.

This is the basic idea of **confounding**. A confounder is a variable that is related both to treatment assignment and to the outcome. Prior earnings are a natural example in this setting.

A natural first thing to do is to control for the observed pre-treatment covariates in a regression model. For example, we can fit

\[
\texttt{re78} = \beta_0 + \beta_1 \texttt{treat} + \beta_2 X + \varepsilon,
\]

where $X$ denotes the baseline covariates.

```{r}
mod_adj <- lm(
  re78 ~ treat + age + educ + race + married + nodegree + re74 + re75,
  data = aa
)

mod_adj |>
  tidy(conf.int = TRUE)
```

The treatment coefficient changes noticeably after adjustment. Now, it is statistically significant and **positive**. Our estimate is that the job training program is associated with an increase of 1548 dollars after adjusting for the other covariates.

Regression adjustment is therefore useful. It can remove bias due to **observed** baseline differences, at least when the model is doing a reasonable job of adjustment. At the same time, it is important not to say more than the model justifies. It would **not** be justified to say that participation in the pogram **increased** 1978 earnings, because that is assigning causality to the program. All we know is that it is **associated** with an increase after adjusting for the other covariates. In order to justify a causality claim, we would need to believe that after controlling for the observed pre-treatment covariates, the treated and untreated groups are comparable in all ways relevant to the outcome.

This is a much stronger claim than simply saying that we included several variables in a regression. A regression coefficient is not automatically causal just because it comes from a multiple regression model. The causal interpretation comes from the assumptions we are willing to defend.

A good way to summarize the distinction is the following:

- The naive model compares treated and untreated units directly.
- The adjusted model compares treated and untreated units after controlling for observed baseline covariates.
- A causal interpretation requires that this adjusted comparison be a good stand-in for the missing *counterfactual* comparison, see below for details on counterfactuals.

One way to formalize the causal question is through the potential outcomes framework. For each individual $i$, let

- $Y_i(1)$ be the earnings that person $i$ would have if treated;
- $Y_i(0)$ be the earnings that person $i$ would have if untreated.

The individual causal effect is then

\[
Y_i(1) - Y_i(0).
\]

The difficulty is that we never observe both values for the same person. If person $i$ is treated, then we observe $Y_i(1)$ but not $Y_i(0)$. If person $i$ is untreated, then we observe $Y_i(0)$ but not $Y_i(1)$. The observed outcome can be written as

\[
Y_i = D_iY_i(1) + (1-D_i)Y_i(0),
\]

where $D_i$ is the treatment indicator. That is, $D_i = 1$ if treatment is received and $D_i = 0$ if no treatment is received.

This is the fundamental problem of causal inference. For each person, one of the two relevant outcomes is missing. In observational data, we use untreated individuals as stand-ins for the missing untreated outcomes of the treated group, or treated individuals as stand-ins for the missing treated outcomes of the untreated group. The central question is whether those stand-ins are credible.

In the present example, the raw comparison is not especially credible because the groups differ so much at baseline. The adjusted regression is more credible, but only to the extent that the included covariates capture the important differences between the two groups. In particular, there is nothing about the multiple regression that directly tries to make the two groups comparable; if it happens, it is fortuitous. 

Later, we will study a different way to organize this comparability problem. Rather than putting all of the adjustment into a single outcome regression, we can first ask how likely each unit was to receive treatment based on the observed pre-treatment covariates.

This leads to the **propensity score**, defined by

\[
e(X) = P(D = 1 \mid X).
\]

Here, $X$ denotes the observed pre-treatment covariates. Units with similar propensity scores have similar observed baseline characteristics, at least in the sense relevant for treatment assignment. The idea is then to compare treated and untreated units that were similarly likely to receive treatment in the first place.

We will not develop propensity scores fully in this section, but we can at least estimate them and inspect the results. A common first step is to fit a logistic regression for treatment.

```{r}
ps_mod <- glm(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = aa,
  family = binomial()
)

aa_ps <- ps_mod |>
  augment(data = aa, type.predict = "response") |>
  mutate(
    treat = factor(treat, levels = c(0, 1), labels = c("Control", "Treated"))
  )

ps_mod |>
  tidy()
```

We can now look at the estimated propensity scores by treatment group.

```{r}
aa_ps |> 
  ggplot(aes(x = .fitted, fill = treat)) +
  geom_histogram(alpha = 0.5, bins = 30, position = "identity") +
  labs(x = "Estimated propensity score", y = "Count", fill = NULL)
```

The plot is not itself a causal analysis, but it begins to show us something that the outcome regression did not show directly: how much overlap there is between the treated and untreated groups on the observed baseline covariates. Later, this idea of overlap will become central. In this particular plot, notice that there is not a lot of overlap between control and treated in the low propensity score group. This means that there is some combination of covariates which almost uniformly led participants not to get job training. That is somewhat bad news for causal analysis in general, and certainly should lead to hesitancy in interpreting the multiple regression output causally. We expect to see some predictive power of the covariates on whether the treatment was received, but we are hoping to still see significant overlap for our causal analysis. There are techniques for dealing with data like this which we will discuss in later sections.

### Summary

This section makes three points.

First, a simple regression of outcome on treatment can be misleading in observational data because treated and untreated groups often differ before treatment occurs.

Second, regression adjustment can account for observed baseline differences and can change the treatment estimate substantially. In the job-training example, this is exactly what happens.

Third, even an adjusted regression is not automatically causal. The causal interpretation requires additional assumptions about comparability between the treated and untreated groups after adjustment.

Propensity scores give us another way to think about that comparability. They do not remove the need for assumptions, but they make the adjustment problem more explicit. In the next part of our causal inference discussion, we will use them to study matching and related methods.

## Identification Assumptions for Causal Adjustment

In the previous section, we saw that a naive regression of later earnings on treatment in the Lalonde data gave a different answer than a regression that adjusted for observed baseline covariates. That already tells us something important: the treated and control groups were different before treatment. In observational work, that is the rule rather than the exception.

The next question is the more important one. When does an adjusted comparison deserve a causal interpretation? Regression adjustment can be part of a causal analysis, but the causal interpretation does not come from the software output. It comes from the assumptions that justify using treated and untreated units as stand-ins for one another's missing counterfactual outcomes.

Throughout this section, we continue to use the Lalonde job-training data. The treatment is participation in the job-training program, and the outcome is earnings in 1978.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(MatchIt)
library(broom)

data("lalonde")


aa <- lalonde |>
  as_tibble() |>
  mutate(
    treat = factor(treat, levels = c(0, 1), labels = c("Control", "Treated")),
    race = factor(race),
    married = factor(married, levels = c(0, 1), labels = c("No", "Yes")),
    nodegree = factor(nodegree, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

We will use the following observed pre-treatment covariates throughout this section:

- `age`
- `educ`
- `race``
- `married`
- `nodegree`
- `re74`
- `re75`

### Assumptions for Causal Adjustment

We continue to write the potential outcomes for person $i$ as $Y_i(1)$ and $Y_i(0)$, where these represent that person's 1978 earnings under treatment and under control, respectively. The observed outcome is

$$
Y_i = D_iY_i(1) + (1-D_i)Y_i(0).
$$

This is the formal statement of **consistency**. If a person is treated, then the observed outcome is the treated potential outcome; if a person is untreated, then the observed outcome is the untreated potential outcome. This sounds automatic, but it depends on the treatment being defined clearly enough that $Y(1)$ and $Y(0)$ are meaningful. If the treatment were simply coded as "exercise," then the treated condition might bundle together walking, weight lifting, marathon training, yoga, and many other interventions. In that situation, the meaning of $Y(1)$ becomes murky. Likewise, if "job training" bundled together many radically different programs, then the treated potential outcome would be less clearly defined.

The most important assumption in observational work is **ignorability**:

$$
(Y(1), Y(0)) \perp D \mid X.
$$

In words, after conditioning on the observed pre-treatment covariates $X$, treatment assignment is independent of the potential outcomes. A slightly more informal version is often easier to remember: once we control for the relevant observed pre-treatment variables, treated and untreated units are comparable on average. This assumption is also called **no unmeasured confounding** or **selection on observables**.

It is important to be clear about what ignorability does and does not say. It does not say that treatment is random in the raw data. It says that treatment is *as if random after conditioning on the chosen covariates*. That is why the choice of covariates matters so much.

This connects directly to the informal **backdoor idea**. If a variable is a common cause of treatment and outcome, then failing to adjust for it creates a noncausal association between treatment and outcome. In the Lalonde data, prior earnings are a natural example. Someone's 1975 earnings may affect whether they enter the training program, and 1975 earnings will also help predict 1978 earnings. In that case, `re75` is a confounder, and adjusting for it blocks a noncausal path between treatment and outcome.

A third important assumption is **positivity**, sometimes called **overlap**:

$$
0 < P(D = 1 \mid X = x) < 1
$$

for the values of $x$ we care about. In plain language, this means that for covariate profiles of interest, there must be some chance of both treatment and control. If units with certain covariates are always treated, or always untreated, then we do not have comparable treated and control units in that part of the covariate space. In practice, this means that causal adjustment requires overlap. No overlap means no apples-to-apples comparison.

A final assumption that is worth stating **no interference**. One person's treatment should not directly affect another person's outcome. In the job-training setting, this could fail if treated and untreated workers are competing for the same jobs in the same market. We will not focus on interference in this chapter, but it is worth keeping in mind.

These assumptions help explain why regression is not the opposite of causal inference. Regression is one tool that causal inference can use. The causal interpretation depends on whether the regression is implementing a defensible comparison.

There is one more practical point to make before returning to the Lalonde data. A good control is generally a variable measured before treatment that affects both treatment assignment and the outcome. A bad control is a variable that is downstream of treatment or a collider. For example, if a job-search intervention improves later mental health partly by increasing job-search confidence, then controlling for job-search confidence would remove part of the very treatment effect we are trying to estimate. If the goal is the total effect of treatment, post-treatment mediators are bad controls.

The `jobs` data in the `mediation` package give a nice example. The treatment variable is `treat`, the outcome is `depress2`, and `job_seek` is a plausible post-treatment mediator. In the code below, the third model adds `job_seek` after treatment has already occurred.

```{r}

jobs2 <- mediation::jobs |>
  as_tibble()

mod1 <- lm(depress2 ~ treat, data = jobs2)
mod2 <- lm(depress2 ~ treat + econ_hard + sex + age, data = jobs2)
mod3 <- lm(depress2 ~ treat + econ_hard + sex + age + job_seek, data = jobs2)

bind_rows(
  tidy(mod1) |> mutate(model = "Treatment only"),
  tidy(mod2) |> mutate(model = "Baseline controls"),
  tidy(mod3) |> mutate(model = "Adds post-treatment variable")
) |> 
  filter(term == "treat") |> 
  dplyr::select(model, estimate, std.error, statistic, p.value)
```

When the coefficient on treatment changes after `job_seek` is added, that is not automatically an improvement. If `job_seek` lies on the causal pathway from treatment to the outcome, then controlling for it changes the estimand. This is why "control for everything" is not safe advice.

### The Lalonde Data Revisited

We return now to the Lalonde data. The first thing to notice is that the treated and control groups differ on several observed pre-treatment covariates.

```{r}
aa |>
  group_by(treat) |>
  summarize(
    n = n(),
    age = mean(age),
    educ = mean(educ),
    re74 = mean(re74),
    re75 = mean(re75),
    .groups = "drop"
  )
```

```{r}
aa |>
  group_by(treat) |>
  summarize(
    percent_black = mean(race == "black"),
    percent_hispanic = mean(race == "hispan"),
    percent_white = mean(race == "white"),
    married = mean(married == "Yes"),
    nodegree = mean(nodegree == "Yes"),
    .groups = "drop"
  )
```

These are all pre-treatment variables, so differences here are a warning sign for confounding rather than an effect of treatment. In particular, prior earnings differ between the groups, and prior earnings are highly relevant for later earnings.

We can see again that the naive and adjusted regressions tell different stories.

```{r}
mod_naive <- lm(re78 ~ treat, data = aa)

mod_adj <- lm(
  re78 ~ treat + age + educ + race + married + nodegree + re74 + re75,
  data = aa
)

bind_rows(
  tidy(mod_naive, conf.int = TRUE) |> mutate(model = "Naive"),
  tidy(mod_adj, conf.int = TRUE) |> mutate(model = "Adjusted")
) |>
  filter(term == "treatTreated") |>
  dplyr::select(model, estimate, conf.low, conf.high, p.value)
```

The sign change is a strong indication that the raw treatment comparison was confounded by observed baseline differences. This does not prove that the adjusted model is causal, but it does show that adjustment matters.

It is useful to visualize overlap directly. We start with one important pre-treatment covariate, 1975 earnings.

```{r fig.height=4}
ggplot(aa, aes(x = re75, fill = treat)) +
  geom_density(alpha = 0.4) +
  labs(x = "1975 earnings", y = "Density", fill = NULL)
```

Where the two distributions overlap, comparison is more plausible. Where they are far apart, adjustment begins to rely on extrapolation. This same idea can be summarized more compactly by the **propensity score**, which we define as

$$
e(X) = P(D = 1 \mid X).
$$

For now, we use the propensity score only as a way to study overlap. We estimate it with logistic regression using the observed pre-treatment covariates.

```{r}
ps_mod <- glm(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = aa,
  family = binomial()
)

aa_ps <- ps_mod |>
  augment(data = aa, type.predict = "response") |>
  mutate(ps = .fitted) |>
  dplyr::select(-.fitted)
```

Now we look at the distributions of the estimated propensity scores in the treated and control groups.

```{r fig.height=4}
ggplot(aa_ps, aes(x = ps, fill = treat)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 20) +
  labs(x = "Estimated propensity score", y = "Count", fill = NULL)
```

The treated units tend to have higher estimated probabilities of treatment than the control units. This is the expected direction. The issue is not that the graph is backwards. The issue is that the groups may be too separated. Good causal adjustment needs overlap, not simply a treatment model that predicts treatment well. If treated and control units have little overlap in propensity score, then there are few comparable units to compare.

This is the main reason to introduce propensity scores in the first place. Regression adjustment hides the comparability problem inside a single coefficient. Propensity scores make it easier to inspect whether treated and untreated units are similar on the observed pre-treatment covariates and whether common support is adequate.

The big lesson from this section is that adjusted comparisons only deserve a causal interpretation under assumptions. Consistency tells us that the treatment must be defined clearly enough for the potential outcomes to make sense. Ignorability says that after conditioning on the observed pre-treatment covariates, treatment assignment is as good as random. Positivity says that comparable treated and control units must exist. Finally, the discussion of bad controls reminds us that adjustment can go wrong when we condition on variables that are affected by treatment. The next section uses propensity scores more directly to organize causal comparisons in observational data.

### Vocabulary and an Example

In causal inference, we imagine that each unit has more than one possible outcome. For a household in the `retrofit` study, one outcome is the amount of gas it would use if it received the retrofit, and another is the amount it would use if it did not. These are called the **potential outcomes**. We never observe both outcomes for the same household, which is why causal inference is difficult in the first place. The whole enterprise is an attempt to learn something about the missing counterfactual outcome from the data we do observe.

**Consistency** says that the observed outcome under the treatment actually received is the corresponding potential outcome. If a household received the retrofit, then its observed gas use is its treated potential outcome; if it did not, then its observed gas use is its untreated potential outcome. This sounds automatic, but it depends on the treatment being defined clearly enough that the phrase “received the retrofit” has a reasonably stable meaning. If treated households received wildly different interventions, then the treated potential outcome $Y(1)$ would be less clearly defined.

**Ignorability** means that after we condition on the observed pre-treatment covariates, treatment assignment is independent of the potential outcomes. In practice, this is the assumption that after adjusting for the right observed baseline variables, treated and untreated units are comparable on average. This is also called **no unmeasured confounding** or **selection on observables**. It is the key assumption that turns an adjusted observational comparison into a candidate causal estimate.

A **confounder** is a variable that affects both treatment assignment and the outcome. Confounders are the reason naive treated-versus-control comparisons can be misleading. A good confounder to keep in mind is a variable that helps explain both who receives treatment and what the outcome would have looked like anyway. In the `retrofit` example, prior gas use is a natural candidate because households with high prior gas use may be more likely to receive the retrofit and also more likely to have high future gas use.

The informal **backdoor idea** is a way to think about confounding without too much notation. If a variable is a common cause of treatment and outcome, then it creates a noncausal path between them. Adjusting for that variable blocks that path. This gives a practical way to think about which variables belong in an adjustment set: we look for pre-treatment variables that help explain why treated and untreated units differ before treatment begins.

**Positivity**, also called **overlap**, means that for the covariate values we care about, there must be some chance of both treatment and control. In plain language, the treated and untreated groups must actually contain comparable units. Even if we measure many sensible covariates, a causal analysis becomes difficult if some kinds of units are almost always treated and others are almost always untreated. Lack of overlap is not something that adjustment can solve; it is a limitation of the data.

A **propensity score** is the probability of treatment given the observed pre-treatment covariates. It summarizes many baseline covariates into a single number that describes how likely treatment was, based on what we observed before treatment began. Propensity scores do not remove unmeasured confounding, but they are useful for checking overlap and organizing comparisons among units that look similar on the observed covariates.

A **mediator** is a variable that lies on the causal pathway from treatment to outcome. Mediators are important if we want to understand *how* treatment works, but they are usually bad controls if the goal is the **total effect** of treatment. Controlling for a mediator blocks part of the very effect we are trying to estimate.

A **collider** is a variable that is caused by two other variables. Conditioning on a collider can create a spurious association between those causes, which is why colliders are bad controls. This is different from the mediator problem. A mediator is bad because it blocks part of a real causal effect. A collider is bad because conditioning on it can open a false path and induce bias. Note that often a collider is caused by one variable in the data (such as treatment) and one unmeasured variable (such as organization or motivation). Colliders can be hard to see, but it is important not to condition on them.

A **bad control** is a variable that should not be conditioned on in a causal analysis. Two common kinds of bad controls are mediators and colliders. The main lesson is that “control for everything” is not a safe rule. Some variables help create comparability; others distort the comparison.

One last idea is that of **no interference**: one unit’s treatment should not directly affect another unit’s outcome. In the `retrofit` study, this would mean that one household’s retrofit does not directly change another household’s gas use.

# The `retrofit` study

The `retrofit` study is a hypothetical observational study of a city-run home weatherization program. The city offers subsidies for attic and wall insulation, air sealing, duct repair, and minor furnace improvements to households with high energy burdens. Some households apply directly, while others are identified through neighborhood outreach or energy audits. Some households receive the retrofit before the winter heating season, and some do not. There is no random assignment. Participation depends on a mixture of household choice, city targeting, contractor availability, and administrative prioritization.

The treatment variable is `retrofit`, equal to 1 if the household received the weatherization package before winter 2024--25 and 0 otherwise. The outcome is `gas_use_2025`, the household’s total natural-gas use during winter 2024--25. Lower values are better. Because treatment is not randomized, treated and untreated households are likely to differ in many ways before treatment begins. Some live in older homes, some have larger families, some have higher energy burdens, and some have homes that are much less energy efficient than others. Those pre-treatment differences are exactly what make causal adjustment necessary.

It is useful to imagine what `?retrofit` might look like in R.

```text
retrofit {causalexamples} R Documentation

Home Weatherization Program Study

Description
Observational study of a city-run home weatherization program. The dataset follows
5,000 owner-occupied households over two winters. Treated households received a
subsidized weatherization package before winter 2024--25. The main outcome is
natural-gas use during winter 2024--25. The data were assembled from utility bills,
program records, property records, and a follow-up survey.

Usage
retrofit

Format
A data frame with 5000 rows and 27 variables:

retrofit
    Treatment indicator equal to 1 if the household received the retrofit package
    before winter 2024--25 and 0 otherwise.

gas_use_2025
    Outcome: total natural-gas use during winter 2024--25, measured in therms.

gas_use_2024
    Baseline variable measured before treatment: total natural-gas use during
    winter 2023--24, measured in therms.

electric_use_2024
    Baseline variable measured before treatment: total electricity use during
    calendar year 2024 before retrofit work began, measured in kWh.

sqft
    Baseline variable measured before treatment: interior square footage of the
    home from property records.

home_age
    Baseline variable measured before treatment: age of the home in years at the
    start of winter 2024--25.

building_type
    Baseline variable measured before treatment: detached, duplex, rowhouse, or
    small multifamily.

insulation_baseline
    Baseline variable measured before treatment: insulation score from the
    pre-treatment energy audit.

furnace_age
    Baseline variable measured before treatment: age of the primary heating system
    in years at intake.

income
    Baseline variable measured before treatment: household income in thousands of
    dollars, recorded at intake.

arrears_balance
    Baseline variable measured before treatment: amount owed on utility bills at
    intake or referral.

credit_score
    Baseline variable measured before treatment: household credit-score category
    at intake, before any retrofit work begins.

owner_occupied
    Baseline variable measured before treatment: indicator for owner occupancy at
    intake.

years_in_home
    Baseline variable measured before treatment: number of years the household had
    lived in the home at intake.

occupants
    Baseline variable measured before treatment: number of people living in the
    home at intake.

children
    Baseline variable measured before treatment: number of children in the
    household at intake.

elderly_resident
    Baseline variable measured before treatment: indicator for whether at least
    one resident was age 65 or older at intake.

work_from_home_days
    Baseline variable measured before treatment: average number of days per week
    at least one adult worked from home at intake.

energy_audit_score
    Baseline variable measured before treatment: audit score summarizing energy
    inefficiency before treatment.

energy_burden
    Baseline variable measured before treatment: share of household income spent
    on home energy before treatment.

neighborhood
    Baseline variable measured before treatment: neighborhood identifier from the
    city's administrative records.

application_month
    Baseline variable measured before treatment: month in 2024 when the household
    applied or was referred to the program.

insulation_added
    Post-treatment variable: amount of new insulation installed after treatment
    assignment as part of the retrofit intervention.

contractor_visits
    Post-treatment variable: number of contractor visits after treatment began.

coaching_sessions
    Post-treatment variable: number of energy coaching sessions completed after
    treatment began.

thermostat_setting_2025
    Post-treatment variable: average winter thermostat setting during 2024--25.

inspection_passed
    Post-treatment variable: indicator for whether the household passed the
    post-installation inspection.

followup_survey_response
    Post-treatment variable: indicator for whether the household completed the
    spring 2025 follow-up survey.
```


### Worksheet

After you read the vocabulary and the `retrofit` description, switch to the worksheet. The goal is not to produce the one perfect answer. The goal is to practice naming the roles variables can play in an observational causal study and to justify those choices in words. 

Consult the reference handout while you work. First answer the questions on your own. Then compare answers with a partner or small group.

#### Part A: Core roles

1. What is the treatment variable?

2. What is the outcome variable?

#### Part B: Variable classification

Fill in each line with several variable names (unless otherwise instructed) from the `retrofit` study.

- **Strong pre-treatment confounders:**

- **Other useful pre-treatment covariates:**

- **One post-treatment mediator (best choice):**

- **One possible collider (best choice):**

- **Other bad controls (not including your mediator or collider choice):**

- **Variables or combinations of variables that raise an overlap concern:**

- **Variables that raise a consistency concern:**

#### Part C: One step back

In two or three sentences, explain the difference between these two problems:

- unmeasured confounding
- limited overlap

Your answer should make clear why one is something we try to reduce through adjustment, while the other is a limitation of the data themselves.

### Solutions

This key is not completely unique. Several answers are defensible if the explanation is good. The main goal is that students classify variables according to their causal role and justify the harder cases clearly.

#### Part A: Core roles

1. **Treatment variable:** `retrofit`

2. **Outcome variable:** `gas_use_2025`

#### Part B: Variable classification

- **Strong pre-treatment confounders:**  
  `gas_use_2024`, `sqft`, `home_age`, `insulation_baseline`, `furnace_age`, `income`, `arrears_balance`, `energy_burden`, `energy_audit_score`, `neighborhood`

- **Other useful pre-treatment covariates:**  
  `electric_use_2024`, `building_type`, `owner_occupied`, `years_in_home`, `occupants`, `children`, `elderly_resident`, `work_from_home_days`, `credit_score`, `application_month`

- **One post-treatment mediator (best choice):**  
  `thermostat_setting_2025`

- **One possible collider (best choice):**  
  `followup_survey_response`

- **Other bad controls (not including your mediator or collider choice):**  
  `coaching_sessions`, `contractor_visits`  
  A reasonable additional answer is `inspection_passed` if students explain that conditioning on it may induce selection bias or open a noncausal path. I would usually not count `insulation_added` here if the goal is to keep the roles clean, since it is more useful for the consistency discussion.

- **Variables or combinations of variables that raise an overlap concern:**  
  Extreme combinations are the main idea here. Good answers include things like:
  
  - very high `income` together with low `energy_burden`
  - very new homes with strong `insulation_baseline`
  - very old homes with high `gas_use_2024`, poor `insulation_baseline`, and high `energy_burden`
  - combinations of `neighborhood`, `income`, and `home_age`
  
  The point is that some treated and untreated households may be scarce in these parts of the covariate space.

- **Variables that raise a consistency concern:**  
  `insulation_added` is the best single answer. Stronger answers may note that `insulation_added` together with `contractor_visits` or `coaching_sessions` suggests that `retrofit = 1` may bundle together substantially different interventions.

#### Part C: One step back

A good answer would say something like this:

> Unmeasured confounding means that there may still be variables affecting both treatment assignment and the outcome that are not included in the data, so adjustment may fail to make treated and control units truly comparable. Limited overlap is different: it means that even after choosing the covariates, there may be parts of the covariate space where treated and control units are too different to compare directly. Adjustment is meant to reduce confounding by observed variables, but it cannot repair a lack of overlap, which is a limitation of the data themselves.

## Collid3rs

A variable can be a **collider** without automatically being a bad control for the causal effect of interest. A collider is simply a variable with two arrows pointing into it on some path. For example, if both `home_age` and `furnace_age` affect `energy_burden`, then `energy_burden` is a collider relative to those two variables. That fact alone does **not** imply that controlling for `energy_burden` will bias the estimated effect of `retrofit` on `gas_use_2025`.

What matters for causal inference is whether conditioning on the collider opens a **noncausal path between treatment and outcome**. This is the key distinction. A variable may be locally a collider, yet still be harmless for the treatment effect if conditioning on it does not create a new path from `retrofit` to `gas_use_2025`. In that case the variable might still cause ordinary modeling problems such as redundancy, multicollinearity, or interpretability issues, but not collider bias in the treatment effect.

A collider is **not** automatically a bad control. It becomes a bad control when conditioning on it opens a spurious path from treatment to outcome.

A good way to separate the ideas is to compare two cases.

In a relatively harmless case, suppose

\[
\text{home\_age} \rightarrow \text{energy\_burden} \leftarrow \text{furnace\_age}.
\]

Here `energy_burden` is indeed a collider of `home_age` and `furnace_age`. But unless conditioning on `energy_burden` opens some new route from `retrofit` to `gas_use_2025`, this local collider structure does not itself create causal bias in the treatment effect. The issue is mostly statistical rather than causal.

In a harmful case, suppose a pre-treatment variable such as `finalist_pool` behaves like this:

\[
\text{retrofit} \leftarrow U_1 \rightarrow \text{finalist\_pool} \leftarrow U_2 \rightarrow \text{gas\_use\_2025},
\]

where `U_1` is an unmeasured factor such as household persistence in the application process, and `U_2` is an unmeasured factor such as hidden housing damage. If we condition on `finalist_pool`, then treatment can become associated with hidden housing damage through the opened path. Since hidden housing damage affects later gas use, the estimated treatment effect becomes biased. This is the classic collider problem.

One important intuition is that before conditioning on the collider, `U_2` is not associated with treatment along that path because the path is blocked. After conditioning on the collider, `U_2` may become associated with treatment. That is why conditioning on a collider can create a confounding-like problem that was not there before.

## Post-treatment variables do not always fit neatly into one box

Some bad controls do not fit neatly into the simple mediator-versus-collider classification. A good example is `sqft` measured **after** treatment. Baseline square footage is a sensible pre-treatment covariate. But if post-treatment `sqft` reflects later additions to the home, then it is no longer just a baseline housing characteristic. Instead, it may reflect changes that happened after treatment or later household decisions influenced by treatment.

That means post-treatment `sqft` is usually best described as a **messy post-treatment bad control**. It is not the cleanest flagship example of a mediator, and it is not naturally a collider either. The main problem is simply that it is measured after treatment and may itself reflect downstream changes. A useful sentence is: not every bad control falls neatly into the mediator/collider taxonomy; some are problematic simply because they reflect post-treatment developments.

## Variables measured after treatment are not automatically bad controls

The crucial issue is not just whether a variable is measured after treatment in calendar time. The real question is whether the variable is **downstream of the treatment variable in the causal question of interest**. This is why `heating_degree_days` is different from post-treatment square footage.

If `heating_degree_days` measures how cold the winter was, then it is not caused by the retrofit. Even if it is recorded during the same winter as the outcome, it is not downstream of treatment. For that reason, it is not a bad control in the same way a post-treatment mediator would be. It may still or may not be useful, depending on whether households face different weather conditions, but it is not problematic merely because of timing.

So the better rule is this: be very careful with variables measured after treatment, but the real danger comes from variables that are affected by treatment or otherwise lie on the treatment pathway.

## Why `insulation_added` and `thermostat_setting_2025` play different roles

The variable `insulation_added` is best used in the retrofit example to illustrate a **consistency** concern. If treated households receive very different amounts of insulation or very different bundles of work, then the label `retrofit` may hide important versions of treatment. In that case the treated potential outcome is less clearly defined, and consistency becomes less plausible unless treatment is described more precisely.

The variable `thermostat_setting_2025` is a cleaner example of a **mediator**. If the retrofit makes the home easier or cheaper to heat, households may change how warm they keep the house, and that changed thermostat behavior may affect `gas_use_2025`. Then

\[
\text{retrofit} \rightarrow \text{thermostat\_setting\_2025} \rightarrow \text{gas\_use\_2025}
\]

is a plausible part of the causal story. If the goal is to estimate the **total effect** of retrofit on gas use, then controlling for `thermostat_setting_2025` would be a mistake because it would block part of the treatment effect.

## Why `finalist_pool` or `special_review` can be problematic in more than one way

Variables such as `finalist_pool` or `special_review` are not only potential colliders. They can also cause trouble because they often function as **selection variables**. If the analysis is restricted to households in the finalist pool, then the study is no longer describing the original applicant population. Even apart from collider bias, this restriction changes the population being studied.

These variables may also be poor modeling choices because they are often **opaque composite variables**. A status such as `finalist_pool` may bundle together income, energy burden, application quality, housing problems, and city rules into one summary indicator. Even if it were not creating collider bias, it could still be unhelpful because it hides the underlying sources of variation rather than representing a clean pre-treatment characteristic.

So there are three distinct concerns worth separating:

- **Collider problem:** conditioning on the variable may open a false treatment--outcome path.
- **Selection problem:** restricting the sample using the variable may change the target population.
- **Modeling problem:** the variable may bundle together several processes and be hard to interpret causally.

## A compact summary

The retrofit discussion suggests four practical lessons.

1. A variable can be a collider somewhere in the graph without being a bad control for the treatment effect.
2. A collider becomes a causal problem when conditioning on it opens a noncausal path from treatment to outcome.
3. Not every bad control is a clean mediator or collider; some post-treatment variables are simply messy because they reflect downstream changes.
4. Variables such as `finalist_pool` can be problematic for more than one reason: collider bias, selection bias, and opacity in the adjustment set.

These distinctions are worth making slowly, because people often hear "collider" or "post-treatment variable" as if those labels automatically settle the matter. The more careful rule is always the same: ask whether conditioning on the variable helps or harms the treatment--outcome comparison that the causal question actually requires.



## Directed Acyclic Graphs (DAGs)

DAGs can be used to help understand the relationship between variables in a causal analysis. In its simplest form, we put an arrow bewteen two variables if there is or could be (we believe) a causal effect between the two variables. As a basic example, if $D$ is the 0/1 treatment variable and $R$ is the response variable, then we would normally put an arrow between $D$ and $R$. In the context of the `lalonde` data, we believe that there is or could be a causal effect between `treat` and `re78`, so we would write

\[
\text{treat} \to \text{re78}
\]

We also believe that `re75` has a causal effect on `re78`, so we would write

\[
\text{re75}\to \text{re78}
\]

At this point, `re75` is not yet identified as a confounder, merely a variable that could have a causal effect on `re78`. However, we also believe that income in 1975 could have a causal effect on whether a candidate participates in the job training program, so we would have:

\[
\text{treat} \leftarrow \text{re75}\rightarrow \text{re78}
\]
This is the classic DAG for a confounder. It is a variable that has a causal effect on both treatment and the response variable. This creates a **backdoor path** so that if we do not adjust for income in 1975, then we could have non-causal correlation between treatment and income in 1978. In other words, `re75` is a good control because it helps block a noncausal association between `treat` and `re78`.

Now let's look at **mediators**. Suppose we also had the variable `jobsearcheffort`. This would be a mediator. The reason is that we can imagine that the training could have a causal effect on the job search effort. Job search effort also has a causal effect on income. So we get the DAG

\[
\text{treat} \rightarrow \text{jobsearch} \rightarrow \text{re78}
\]

This is the traditional mediator DAG. If we control on `jobsearch`, then part of the effect of `treat` will be swallowed up by the `jobsearch` estimate, when in fact the `jobsearch` usage was **caused** by the `treat`. Note that the difference between a confounder and a mediator is whether the variable is caused by the treatment (mediator) or causes the treatment (confounder). 

Finally, let’s look at **colliders**. We claim that `followupresponse` is a collider. The causal story is that the training can affect whether someone completes the follow-up, and organizational skill (which is unmeasured) can also affect whether someone follows up. Organizational skill may also affect later income. This gives the DAG

\[
\text{treat} \rightarrow \text{followupresponse} \leftarrow \text{organization} \rightarrow \text{re78}.
\]

By itself, this path isn't a problem. But if we condition on `followupresponse` for example, by controlling for it in a regression or by restricting attention to those who respond, we open a noncausal path between treatment and income. Treatment does not affect real income along this path, but conditioning on the collider can still induce a correlation between treatment and income.

Although many colliders which are bad controls are downstream of treatment, that is not always the case. Here is a slightly more complicated collider example in which the collider is **not** downstream of treatment. Suppose `retrofit` is our treatment and `gas\_use\_2025` is the outcome. Now imagine two unmeasured variables. The first is household persistence: some households are better at completing paperwork, meeting deadlines, and following through with the application process. That can increase the chance that they ultimately receive the retrofit. The second is hidden housing damage: some homes have serious but partly unrecorded problems, such as leaky ducts or damaged window seals, and those problems can increase later gas use. Suppose also that a household enters the city's `finalist\_pool` either because it has a very persistent application or because the home appears complicated enough to merit special attention. Then we have the DAG

\[
\text{retrofit} \leftarrow \text{persistence} \rightarrow \text{finalist\_pool} \leftarrow \text{hidden damage} \rightarrow \text{gas\_use\_2025}.
\]

The key point is that `finalist_pool` is a collider, but it is **not** downstream of treatment. By itself, this path is not a problem. However, if we condition on `finalist_pool`, for example, by controlling for it in a regression or by restricting the analysis only to households in the finalist pool, we open a noncausal path between treatment and outcome. Intuitively, once we hold finalist status fixed, treatment begins to carry information about hidden housing damage through the induced association created at the collider. That means `finalist_pool` is a bad control even though it is not caused by treatment.

We contrast that with the following collider which is not a bad control. This is different from the earlier `energy_burden` example. There, the idea was that `energy_burden` might be a collider relative to two other baseline housing variables, such as `home_age` and `furnace_age`, if both of those affect energy burden. But that fact alone does not make `energy_burden` a bad control for estimating the effect of `retrofit` on `gas_use_2025`. The reason is that conditioning on `energy_burden` does not necessarily open any important new noncausal path from treatment to outcome. In that case, the main concerns are more ordinary modeling issues, such as redundancy with other baseline controls or difficulty interpreting coefficients, rather than collider bias in the treatment effect. However, if you could come up with a different story about how `energy_burden` collides with variables to create a noncausal path between treatment and response, then it could be a bad control.

## Matching

## Propensity scores and matching in the Lalonde data

We now use propensity scores and nearest-neighbor matching to estimate the effect of participation in the training program on `re78`, real earnings in 1978. The goal is to compare treated and untreated individuals who look similar on the observed pre-treatment covariates. This is useful because the training and control groups differ in important baseline ways, so a raw comparison of 1978 earnings is not automatically credible as a causal estimate.

The basic idea is simple. First, we estimate each individual's **propensity score**, which is the probability of treatment given the observed baseline covariates. Then we use those scores to match treated and control individuals with similar covariate profiles. This does not solve the problem of unmeasured confounding, but it can improve comparability on the observed covariates and reduce reliance on extrapolation.

```{r}
library(tidyverse)
library(MatchIt)
library(broom)

aa <- MatchIt::lalonde |>
  as_tibble() |>
  mutate(
    treat = factor(treat, levels = c(0, 1), labels = c("Control", "Treated")),
    race = factor(race),
    married = factor(married, levels = c(0, 1), labels = c("No", "Yes")),
    nodegree = factor(nodegree, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

We begin by estimating propensity scores using a logistic regression for treatment assignment. The covariates here are all measured before treatment and are plausible common causes of treatment and later earnings.

```{r}
ps_mod <- glm(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = aa,
  family = binomial()
)

aa_ps <- ps_mod |>
  augment(data = aa, type.predict = "response") |>
  mutate(ps = .fitted) |>
  select(-.fitted)
```

We can now inspect the estimated propensity scores. The main issue is **overlap**. Where the treated and control groups overlap, there are plausible treated-versus-control comparisons. Where they do not overlap, the analysis becomes more dependent on extrapolation.

```{r}
aa_ps |>
  ggplot(aes(x = ps, fill = treat)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 20) +
  labs(
    x = "Estimated propensity score",
    y = "Count",
    fill = NULL
  )
```

If the treated group tends to have much higher propensity scores than the control group, that tells us treatment assignment is strongly related to the observed baseline covariates. That is exactly why a naive comparison can be misleading. At the same time, if there is at least some substantial overlap, then matching may allow us to compare treated and control individuals who were similarly likely to receive treatment.

Next, we use nearest-neighbor matching based on the estimated propensity scores. The basic algorithm for this is that we pick a treated unit and find the control that has the nearest propensity score. We then remove the treated and control unit from the data and put them into a new data set. We go to the next treated unit and repeat until either there are no more controls or no more treated units.  

```{r}
m_out <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = aa,
  method = "nearest",
  distance = "glm"
)

summary(m_out)
```

The matching summary is useful because it reports balance before and after matching. The main question is whether the treated and control groups become more similar on the observed covariates after matching. In a good matching analysis, we care more about improved balance than about how well the treatment model predicts treatment. Here, we can see that several variables are not well balanced even after matching. We focus only on the first three colums and we see that while the mean of `age` (for example) is well balanced with 25.8 for treatment and 25.3 for control, the racial balance is poor with 84% of the treated population is black while only 47% of the matched control population is black. For each variable, we want the standardized mean difference to be less than, say 0.2, in absolute value to be well-balanced. Here, we see that the race variables are not well matched, nor is distance, which is the raw propensity score.

Despite these warning signs, we now extract the matched data and estimate the treatment effect in the matched sample.

```{r}
aa_match <- match.data(m_out)

mod_match <- lm(re78 ~ treat, data = aa_match, weights = weights) #not recommended

tidy(mod_match, conf.int = TRUE)
```

This matched regression is much simpler than the earlier adjusted regression because the covariate adjustment has largely been moved into the design stage. After matching, we estimate the effect of treatment by comparing outcomes in the matched sample, where treated and control individuals are intended to be more comparable on the observed baseline covariates. Note that the matched data is now paired, which introduces some dependence structure on the data and puts the usage of `lm` on the paired observations into question. We would want to analyze the matched data set taking into account the paired nature of the data. You already know one way of doing that - paired $t$-tests. These are not central to modern causal inference, but for the sake of time, we use them here.

```{r}
aa_wide <- aa_match |> 
  select(treat, re78, subclass) |> 
  pivot_wider(names_from = treat, values_from = re78) 
  
aa_wide |> 
  summarize(mu_treated = mean(Treated),
            mu_control = mean(Control))
t.test(aa_wide$Treated, aa_wide$Control, paired = TRUE)
```

We see that we have an estimate of 894 dollars higher salary based on the paired data, with a confidence interval of $[-495, 2284]$. This is not statistically significant ($p = .2057$).

Note that our matching was not very good - in particular the racial profile of the treated and untreated groups are different. We can try to fix that by forcing an exact match on race, rather than allowing a dubious close enough match to someone of a different race. However, doint so causes other variables to become poorly balanced.

```{r}
m_out2 <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75, data = aa, method = "nearest", distance = "glm", exact = ~race)

summary(m_out2)
```

This matching is arguably better (based on the distance score), but still does not pass the rule of thumb that standardized differences should be between -0.2 and 0.2. Realistically, it doesn't make a lot of sense to keep trying to get a good match here, when we already know of unmeasured confounders (`sex` for example). As an exercise, compute the estimated effect of treatment among the treated based on this new matching, and see whether it is significant.

```{r eval=FALSE, echo=FALSE}
aa_match <- match.data(m_out2)

aa_wide <- aa_match |> 
  select(treat, re78, subclass) |> 
  pivot_wider(names_from = treat, values_from = re78) 
  
aa_wide |> 
  summarize(mu_treated = mean(Treated),
            mu_control = mean(Control))
t.test(aa_wide$Treated, aa_wide$Control, paired = TRUE)
```


## Bernoulli outcomes

In this section, we consider Bernoulli outcomes. We will use the data set `rhc` in the `causaldef` package. Let's start by examining it.

```{r}
aa <- causaldef::rhc |> 
  as_tibble()
```

The help page doesn't provide a whole lot of information about the variables. Note that for causal inference, understanding exactly what the variables are measuring (and when) is essential. We cannot proceed without a better understanding of the variables. I downloaded the original paper and asked ChatGPT to provide more information on the variables based on the descriptions in the paper. Here is what it gave:

### `rhc`

A dataset containing 5,735 critically ill adult patients from the SUPPORT study. The treatment variable is `swang1`, indicating whether the patient received right heart catheterization (RHC, Swan-Ganz catheterization) within the first 24 hours of ICU care. Common outcome variables include `dth30` (30-day mortality), `death` (longer-term mortality indicator), and `t3d30` (survival time up to 30 days).

#### Format

A tibble with 5,735 rows and 63 variables.

#### Variables

- `X`: row index carried over from the imported data set.
- `cat1`: primary disease category used by the study.
- `cat2`: secondary disease category, when applicable.
- `ca`: cancer status, typically coded as no cancer, localized cancer, or metastatic cancer.
- `sadmdte`: study admission date, stored as an integer date code.
- `dschdte`: hospital discharge date, stored as an integer date code.
- `dthdte`: date of death, stored as an integer date code.
- `lstctdte`: date of last contact, stored as an integer date code.
- `death`: indicator for death by the longer-term follow-up window.
- `cardiohx`: indicator for prior cardiovascular disease history.
- `chfhx`: indicator for prior congestive heart failure history.
- `dementhx`: indicator for prior dementia or related neurologic functional impairment.
- `psychhx`: indicator for prior psychiatric history.
- `chrpulhx`: indicator for prior chronic pulmonary disease history.
- `renalhx`: indicator for prior chronic renal disease history.
- `liverhx`: indicator for prior liver disease history.
- `gibledhx`: indicator for prior gastrointestinal bleeding history.
- `malighx`: indicator for prior malignancy history.
- `immunhx`: indicator for prior immunocompromised status.
- `transhx`: indicator for prior transfer from another hospital or prior major transfer history.
- `amihx`: indicator for prior acute myocardial infarction history.
- `age`: age in years.
- `sex`: sex.
- `edu`: years of education.
- `surv2md1`: model-based estimate of the probability of surviving 2 months, computed at day 1.
- `das2d3pc`: Duke Activity Status Index or a closely related baseline functional-status measure.
- `t3d30`: survival time up to 30 days.
- `dth30`: indicator for death within 30 days.
- `aps1`: APACHE III severity score on day 1.
- `scoma1`: Glasgow Coma Score on day 1.
- `meanbp1`: mean blood pressure on day 1.
- `wblc1`: white blood cell count on day 1.
- `hrt1`: heart rate on day 1.
- `resp1`: respiratory rate on day 1.
- `temp1`: body temperature on day 1.
- `pafi1`: ratio of arterial oxygen pressure to fraction of inspired oxygen (PaO2/FiO2) on day 1.
- `alb1`: albumin on day 1.
- `hema1`: hematocrit on day 1.
- `bili1`: bilirubin on day 1.
- `crea1`: creatinine on day 1.
- `sod1`: sodium on day 1.
- `pot1`: potassium on day 1.
- `paco21`: arterial PaCO2 on day 1.
- `ph1`: arterial pH on day 1.
- `swang1`: treatment indicator for receipt of right heart catheterization within the first 24 hours.
- `wtkilo1`: weight in kilograms on day 1.
- `dnr1`: do-not-resuscitate status on day 1.
- `ninsclas`: insurance class.
- `resp`: indicator for a respiratory admission diagnosis.
- `card`: indicator for a cardiovascular admission diagnosis.
- `neuro`: indicator for a neurologic admission diagnosis.
- `gastr`: indicator for a gastrointestinal admission diagnosis.
- `renal`: indicator for a renal admission diagnosis.
- `meta`: indicator for a metabolic admission diagnosis.
- `hema`: indicator for a hematologic admission diagnosis.
- `seps`: indicator for sepsis.
- `trauma`: indicator for trauma.
- `ortho`: indicator for an orthopedic diagnosis.
- `adld3p`: activities-of-daily-living measure recorded later in the study period, typically around day 3.
- `urin1`: urine output on day 1.
- `race`: race category.
- `income`: income category.
- `ptid`: patient identifier.

#### Details

For causal inference, it is useful to think of these variables in broad groups. The treatment is `swang1`. The most common binary outcome is `dth30`, though `death` and `t3d30` are also used. Variables such as `age`, `sex`, `race`, `income`, `cat1`, `cat2`, and the `*hx` variables are the clearest baseline covariates. The day-1 physiology variables, such as `aps1`, `meanbp1`, `hrt1`, `resp1`, and the laboratory measures, are clinically important but raise timing questions because treatment also occurs within the first 24 hours. Variables such as `dth30`, `death`, `t3d30`, and `adld3p` should be treated as outcomes or downstream variables rather than ordinary baseline controls.

#### Source

SUPPORT (Study to Understand Prognoses and Preferences for Outcomes and Risks of Treatments), as used in observational analyses of right heart catheterization in critically ill ICU patients.

OK, let's modify the data a little bit to make sure things are working correctly, then we'll start with propensity scores.

```{r}
aa <- aa |>
  as_tibble() |>
  mutate(
    across(
      c(cardiohx, chfhx, dementhx, psychhx, chrpulhx, renalhx,
        liverhx, gibledhx, malighx, immunhx, transhx, amihx),
      ~ factor(.x, levels = c(0, 1), labels = c("No", "Yes"))
    )
  )
```

We now perform a logistic regression and estimate the overlap. We choose controls of the following type: 

1. General demographics: `age`, `sex`, `race`, `edu`, `income`, and `ninsclas`
2. General health history before treatment: `cat1`, `ca`, `cardiohx`, `chfhx`, `dementhx`, `psychhx`, `chrpulhx`, `renalhx`, `liverhx`, `gibledhx`, `malighx`, `immunhx`, `transhx`, `amihx`
2. Day of treatment health indicator that cannot reasonably be influenced by choice of treatment: `dnr1`

It would be helpful if we knew which of the day of treatment variables were **before** the treatment and which were **after** the treatment, but absent that information we leave out these potential bad controls. We do not include `cat2` because it contains many missing values. If we knew that these were missing becuase that meant there **isn't** a second disease category, then we could replace it with "None" and add it. 

```{r}
mod <- glm(swang1 ~ age + sex + race + edu + income + ninsclas + cat1 + ca +
    cardiohx + chfhx + dementhx + psychhx + chrpulhx + renalhx +
    liverhx + gibledhx + malighx + immunhx + transhx + amihx + dnr1, data = aa, family = binomial)
summary(mod)

broom::augment(mod) |> 
  mutate(propensity = plogis(.fitted)) |> 
  ggplot(aes(x = propensity, fill = swang1)) + 
  geom_histogram(position = "identity", alpha = 0.5, bins = 20) 
```

Let's also count the number of treated and untreated.

```{r}
summary(aa$swang1)
```

Since we considering RHC as the treatment and we will do 1:1 matching of the treated, this looks pretty good. We don't need there to be many RHC treated people on the lower tail, but we do like to see no RHC people in the upper tail, and we do for the most part.

```{r}
match_mod <- matchit(swang1 ~ age + sex + race + edu + income + ninsclas + cat1 + ca +
    cardiohx + chfhx + dementhx + psychhx + chrpulhx + renalhx +
    liverhx + gibledhx + malighx + immunhx + transhx + amihx + dnr1, 
    data = aa,
    method = "nearest",
    distance = "glm")

aa_dat <- match.data(match_mod)

summary(match_mod)
```

We can see that the balance between treated and untreated is better after matching. Now, we want to examine the outcome for the two groups. We consider the `dth30` as our outcome. We note that we have two balanced, paired groups; one corresponding to RHC and one corresponding to no RHC. To analyze this, we first compute the number of concordant and discordant pairs in the paired sample. Let's start by computing the percentage of patients that survived in each treatment.

```{r}
aa_dat |> 
  group_by(swang1) |> 
  summarize(surv = mean(dth30=="No"))
```

Since we have paired data, the natural test is McNemar's test, which is really just a special version of a chi-squared test. We start by counting the discordant and concordant pairs. Pairs are discordant if the outcomes are different and they are concordant if the outcomes are the same.

```{r}
aa_dat |> 
  select(subclass, swang1, dth30) |> 
  pivot_wider(names_from = swang1, values_from = dth30) |> 
  janitor::clean_names() |> 
  mutate(d30_success = case_when(
    rhc == "No" & no_rhc == "No" ~ "both",
    rhc == "Yes" & no_rhc == "No" ~ "no_rhc",
    rhc == "No" & no_rhc == "Yes" ~ "rhc",
    rhc == "Yes" & no_rhc == "Yes" ~ "neither",
    TRUE ~ "uhoh"
  )) |> 
  pull(d30_success) |> 
  table()
```

We see that 987 paired patients were both successes, in 591 of the pairs only the no RHC was successful, in 367 of the pairs only RHC was successful, and 239 of the pairs neither was successful. McNemar's test does a chi-squared test on the items that are **different**, so in other words:

```{r}
chisq.test(c(591, 367), correct = T)
```

There is also a built in R function for this test, which works for more general things:

```{r}
mcnemar.test(matrix(c(987, 591, 367, 239), nrow = 2))
```

It is a good idea to provide an effect size and confidence interval where possible, in order to make the difference between RHC and no RHC more clear. For this case, we recommend the *paired risk difference*, which is computed by

```{r}
(591 - 367)/sum(c(987, 591, 367, 239))
```

This says that there is about a 10 percentage point better chance of success overall using the no RHC technique than the RHC technique. In order to get a simple CI, we can use `t.test` on the raw data:

```{r}
aa_dat |> 
  select(subclass, swang1, dth30) |> 
  pivot_wider(names_from = swang1, values_from = dth30) |> 
  janitor::clean_names() |> 
  mutate(rhc = as.integer(rhc) - 1,
         no_rhc = as.integer(no_rhc) - 1) |> 
  with(t.test(rhc, no_rhc, paired = T))
```

We see that we have a 95 percent CI of 0.075 to 0.13. Let's compare our results to those of the paper. I asked ChatGPT to read the original paper and compare its results to ours, and this is what it said:

Our basic matched analysis and the original paper point in the same direction, but our estimated effect is much larger. In our analysis, after 1:1 nearest-neighbor matching, the estimated 30-day survival rate was 72.3\% for the No RHC group and 62.0\% for the RHC group. Equivalently, this corresponds to about a 10.3 percentage point higher 30-day mortality rate in the RHC group, with a paired risk difference of 0.103 and a 95\% confidence interval from about 0.075 to 0.130. McNemar's test also gave a very small p-value, suggesting a strong difference between the matched groups.

The paper's matched analysis was less dramatic. Connors et al. reported 1008 matched pairs, with 30-day survival of 67.2\% in the No RHC group and 62.5\% in the RHC group. That is still worse for RHC, but the absolute difference is only about 4.7 percentage points rather than the roughly 10 percentage points we found. They reported a matched-pair odds ratio of 1.24 with a 95\% confidence interval from 1.03 to 1.49 and a p-value of .03. So the qualitative conclusion is the same---RHC is associated with worse short-term outcomes---but the paper's estimated effect is smaller and much less statistically overwhelming than the effect from our simplified analysis.

There are several reasons for this difference. First, our propensity score model is much simpler than the one in the paper. We used a conservative set of baseline demographic, diagnostic, comorbidity, and DNR variables, while the paper used a much richer adjustment set that included many additional measures of illness severity and prognosis. In particular, the paper incorporated variables such as APACHE-based severity information, physiologic measurements, laboratory values, functional status, and a model-based estimate of 2-month survival. The point of including those variables was to make the RHC and No RHC patients more comparable in how sick they were at baseline. Since sicker patients were more likely to receive RHC and also more likely to die, leaving out some of those severity variables can make RHC look more harmful than it really is.

Second, the matching strategy in the paper was more selective than the one we used. Our analysis used the default greedy nearest-neighbor matching and kept all 2184 treated patients, pairing each with a control. The paper instead matched patients within disease category and required propensity scores to be very close. As a result, they ended up with only 1008 matched pairs rather than using all treated patients. In other words, the paper was willing to discard many patients in order to create a cleaner comparison. Our analysis uses more data, but it also allows weaker matches, and that can make the estimated treatment effect look larger if the remaining imbalance still favors the No RHC group.

Third, the paper did not rely only on one matched-pair comparison. The authors also fit multivariable outcome models after propensity-score adjustment, looked at survival over multiple time horizons, examined subgroups, and carried out sensitivity analyses asking how large an omitted confounder would have to be to explain away the observed association. These extra analyses do not prove causality, but they do make the overall argument more careful. Our analysis is a good first pass because it shows the basic logic of propensity scores, matching, and paired outcome analysis. The paper's analysis is more sophisticated because it tries harder to address the possibility that the RHC patients were simply sicker in ways not fully captured by a simpler model.

So the right takeaway is not that one analysis is "correct" and the other is "wrong." Rather, the simpler analysis is useful for illustrating the method, while the paper shows what a more serious observational analysis looks like. The fact that both analyses point in the same direction is reassuring. The fact that the paper's estimated effect is smaller is also informative: it suggests that some of the very large difference we saw in the simplified analysis was probably due to the fact that our matching model did not control as fully for severity of illness as the original study did.


