# Identification Assumptions for Causal Adjustment

In the previous section, we saw that a naive regression of later earnings on treatment in the Lalonde data gave a different answer than a regression that adjusted for observed baseline covariates. That already tells us something important: the treated and control groups were different before treatment. In observational work, that is the rule rather than the exception.

The next question is the more important one. When does an adjusted comparison deserve a causal interpretation? Regression adjustment can be part of a causal analysis, but the causal interpretation does not come from the software output. It comes from the assumptions that justify using treated and untreated units as stand-ins for one another's missing counterfactual outcomes.

Throughout this section, we continue to use the Lalonde job-training data. The treatment is participation in the job-training program, and the outcome is earnings in 1978.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(MatchIt)
library(broom)
library(mediation)

data("lalonde")

aa <- lalonde |>
  as_tibble() |>
  mutate(
    treat = factor(treat, levels = c(0, 1), labels = c("Control", "Treated")),
    black = factor(black, levels = c(0, 1), labels = c("No", "Yes")),
    hispan = factor(hispan, levels = c(0, 1), labels = c("No", "Yes")),
    married = factor(married, levels = c(0, 1), labels = c("No", "Yes")),
    nodegree = factor(nodegree, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

We will use the following observed pre-treatment covariates throughout this section:

- `age`
- `educ`
- `black`
- `hispan`
- `married`
- `nodegree`
- `re74`
- `re75`

## Vocabulary and Assumptions

We continue to write the potential outcomes for person $i$ as $Y_i(1)$ and $Y_i(0)$, where these represent that person's 1978 earnings under treatment and under control, respectively. The observed outcome is

$$
Y_i = D_iY_i(1) + (1-D_i)Y_i(0).
$$

**Consistency** is the assumption that the observed outcome under the treatment actually received is equal to the corresponding potential outcome. If a person is treated, then the observed outcome is $Y(1)$; if a person is untreated, then the observed outcome is $Y(0)$. This sounds automatic, but it depends on treatment being defined clearly enough that $Y(1)$ and $Y(0)$ are meaningful quantities. In the Lalonde setting, this means that "job training" should refer to a reasonably coherent intervention rather than to a vague bundle of wildly different programs. In a health study, a treatment coded only as "exercise" would be much harder to interpret, because walking twenty minutes twice a week, marathon training, and a strength program are all very different versions of "exercise." In an education study, a treatment coded as "tutoring" may also be too broad unless we know whether it means one brief session, weekly one-on-one meetings, or a semester-long structured support program.

**Ignorability** is the assumption that after conditioning on the observed pre-treatment covariates $X$, treatment assignment is independent of the potential outcomes:

$$
(Y(1), Y(0)) \perp D \mid X.
$$

This is often described informally by saying that treatment is *as if random after adjustment* for the observed covariates. It is also called **no unmeasured confounding** or **selection on observables**. In the Lalonde example, ignorability would mean that after controlling for age, education, race/ethnicity, marital status, degree status, and prior earnings, whether a person entered the job-training program is no longer related to what their earnings would have been with or without treatment. In a tutoring example, ignorability would mean that once we adjust for prior grades, attendance, and background variables, the remaining tutoring decision is unrelated to the student's unobserved potential exam outcomes. In a medical example, if we compare people who choose a weight-loss medication to people who do not, ignorability would require that after conditioning on observed health history and baseline risk factors, treatment choice is not still being driven by unmeasured variables such as physician judgment, patient motivation, or symptom severity.

A **confounder** is a variable that affects both treatment assignment and the outcome. Confounders are the reason raw treated-versus-control comparisons can be misleading in observational data. In the Lalonde data, prior earnings such as `re75` are natural examples: previous earnings may affect whether someone enters the training program, and they also help predict earnings in 1978. In a study of tutoring and later grades, prior preparation is a confounder because it influences who seeks tutoring and also influences later academic performance. In a study of smoking and health, socioeconomic status can be a confounder because it may affect both smoking behavior and later health outcomes. Confounders are the variables we most want to measure well and adjust for.

The informal **backdoor idea** is a way to think about confounding without too much notation. If a variable is a common cause of treatment and outcome, then it creates a noncausal path between them, and adjusting for that variable blocks that path. In Lalonde, `re75` is a good example: part of the raw association between treatment and later earnings may simply reflect the fact that people with different prior earnings sort differently into treatment. In an education example, prior achievement creates a backdoor path from tutoring to later grades, because the same variable influences both. In a health-policy example, baseline disease severity creates a backdoor path from treatment choice to later outcomes if sicker patients are more likely to receive a more intensive treatment and also have worse outcomes regardless of treatment.

**Positivity**, also called **overlap**, means that for the covariate values we care about, there must be some chance of both treatment and control:

$$
0 < P(D = 1 \mid X = x) < 1.
$$

In plain language, positivity says that treated and untreated units must actually be comparable in the observed data. In the Lalonde setting, if almost no one with very high prior earnings enters the job-training program, then it is hard to learn the treatment effect for people in that part of the covariate space because there are few comparable treated and untreated units. In a housing-subsidy study, if almost all very high-income households are untreated and almost all very low-income households are treated, then comparisons at the extremes rely on extrapolation rather than direct evidence. In a medical example, if a particular therapy is given almost exclusively to the sickest patients, then there may be little overlap between treated and untreated patients with similar baseline risk profiles. Unlike confounding by observed variables, lack of overlap is not something adjustment can solve; it is a feature of the data.

A **propensity score** is the probability of treatment conditional on the observed pre-treatment covariates:

$$
e(X) = P(D = 1 \mid X).
$$

The propensity score is useful because it summarizes many observed covariates into a single number that captures how likely treatment was, given those covariates. In the Lalonde data, a high estimated propensity score means that, based on age, education, marital status, prior earnings, and the other observed baseline variables, a person looked more like someone who would enter the program. In a tutoring study, the propensity score would summarize how likely a student was to attend tutoring based on prior grades, attendance, and background characteristics. In a medical study, the propensity score might summarize how likely a patient was to receive surgery rather than conservative management based on age, comorbidities, and baseline severity. Propensity scores do not eliminate unmeasured confounding, but they are a convenient way to inspect overlap and organize comparisons among units who look similar on the observed covariates.

**SUTVA** is a collection of assumptions about how treatments and outcomes are defined. In this chapter, the most useful new part of SUTVA is **no interference**: one person's treatment should not directly affect another person's outcome. In Lalonde, this could fail if treated and untreated workers are competing for the same jobs in the same local labor market, so that one person's treatment affects another person's employment opportunities. In an education example, no interference could fail if assigning tutoring to some students changes classroom dynamics or teacher attention for other students. In an infectious-disease setting, vaccination is a classic case where no interference is often implausible, because one person's treatment changes the disease risk of others around them.

A **good control** is usually a pre-treatment variable that helps account for differences between treated and untreated units. Good controls are typically related to both treatment assignment and the outcome, which is why they help reduce confounding. In the Lalonde data, `re74`, `re75`, and education are good candidates because they are measured before treatment and are plausibly related both to program participation and later earnings. In a tutoring study, prior grades are usually good controls for the same reason. In a medical study comparing two treatments for heart disease, baseline blood pressure or disease severity would often be good controls because they affect both treatment choice and later outcomes.

A **mediator** is a variable that lies on the causal pathway from treatment to outcome. Mediators are not inherently bad variables, but they are bad controls if the goal is to estimate the **total effect** of treatment, because conditioning on them blocks part of the pathway through which treatment works. In a Lalonde-style story, job training may increase later earnings partly by improving interview skills or job-search intensity; if we controlled for those post-treatment variables, we would be removing part of the treatment effect itself. In a study of tutoring and course performance, the number of hours spent studying after tutoring begins could be a mediator if tutoring increases study time and study time improves grades. In a medical study, blood pressure measured after treatment begins could be a mediator if the treatment lowers blood pressure and lower blood pressure improves later cardiovascular outcomes.

A **collider** is a variable that is caused by two other variables. Conditioning on a collider can create a spurious association between those causes, which is why colliders are bad controls. In a Lalonde-style example, follow-up survey response could be a collider if treatment affects whether someone responds and conscientiousness also affects response, while conscientiousness also affects later earnings; analyzing only survey responders can then induce bias. In an education example, participation in an optional review session could be a collider if tutoring affects review attendance and motivation also affects review attendance, while motivation also affects exam performance. In a medical example, hospital admission can act as a collider if both treatment history and an unmeasured health risk factor affect admission, while the health risk factor also affects the later outcome.

A **bad control** is a variable that should not be conditioned on in a causal analysis because doing so changes the estimand or introduces bias. Two common kinds of bad controls are mediators and colliders. In the Lalonde setting, a post-treatment variable such as job-search intensity after the program begins would be a bad control if the goal is the total effect of treatment on later earnings. In an education setting, controlling for study hours measured after tutoring begins would similarly block part of the tutoring effect. In a medical example, conditioning on a hospital admission variable that is affected both by treatment and by underlying risk can open a spurious path and bias the estimated treatment effect. The main lesson is that "control for everything" is not a safe rule.

The `jobs` data in the `mediation` package give a concrete illustration of a bad control. The treatment variable is `treat`, the outcome is `depress2`, and `job_seek` is a plausible post-treatment mediator. In the code below, the third model adds `job_seek` after treatment has already occurred.

```{r}
data("jobs")

jobs2 <- jobs |>
  as_tibble()

mod1 <- lm(depress2 ~ treat, data = jobs2)
mod2 <- lm(depress2 ~ treat + econ_hard + sex + age, data = jobs2)
mod3 <- lm(depress2 ~ treat + econ_hard + sex + age + job_seek, data = jobs2)

bind_rows(
  tidy(mod1) |> mutate(model = "Treatment only"),
  tidy(mod2) |> mutate(model = "Baseline controls"),
  tidy(mod3) |> mutate(model = "Adds post-treatment variable")
) |>
  filter(term == "treat") |>
  select(model, estimate, std.error, statistic, p.value)
```

When the coefficient on treatment changes after `job_seek` is added, that is not automatically an improvement. If `job_seek` lies on the causal pathway from treatment to the outcome, then controlling for it changes the estimand. This is one reason that causal adjustment requires more thought than ordinary predictive modeling.

## The Lalonde Data Revisited

The vocabulary above can feel abstract if we do not return to the data. In the Lalonde setting, the central problem is that treated and control groups differ on several observed pre-treatment covariates. Those differences are a warning sign for confounding, because they tell us that treatment assignment was related to characteristics that also matter for later earnings.

```{r}
aa |>
  group_by(treat) |>
  summarize(
    n = n(),
    age = mean(age),
    educ = mean(educ),
    re74 = mean(re74),
    re75 = mean(re75),
    .groups = "drop"
  )
```

```{r}
aa |>
  group_by(treat) |>
  summarize(
    black = mean(black == "Yes"),
    hispan = mean(hispan == "Yes"),
    married = mean(married == "Yes"),
    nodegree = mean(nodegree == "Yes"),
    .groups = "drop"
  )
```

In particular, prior earnings differ between the groups, and prior earnings are highly relevant for later earnings. This is exactly the kind of setting in which a naive treated-versus-control comparison is likely to be misleading.

We can see again that the naive and adjusted regressions tell different stories.

```{r}
mod_naive <- lm(re78 ~ treat, data = aa)

mod_adj <- lm(
  re78 ~ treat + age + educ + black + hispan + married + nodegree + re74 + re75,
  data = aa
)

bind_rows(
  tidy(mod_naive, conf.int = TRUE) |> mutate(model = "Naive"),
  tidy(mod_adj, conf.int = TRUE) |> mutate(model = "Adjusted")
) |>
  filter(term == "treatTreated") |>
  select(model, estimate, conf.low, conf.high, p.value)
```

The sign change is a strong indication that the raw treatment comparison was confounded by observed baseline differences. This does not prove that the adjusted model is causal, but it does show that adjustment matters. It also shows why the assumptions introduced above are necessary. The adjusted model only deserves a causal interpretation if the treatment is well defined, the observed covariates are rich enough to make ignorability plausible, and there is enough overlap to support meaningful comparison.

It is useful to visualize overlap directly. We start with one important pre-treatment covariate, 1975 earnings.

```{r fig.height=4}
ggplot(aa, aes(x = re75, fill = treat)) +
  geom_density(alpha = 0.4) +
  labs(x = "1975 earnings", y = "Density", fill = NULL)
```

Where the two distributions overlap, comparison is more plausible. Where they are far apart, adjustment begins to rely on extrapolation. This same idea can be summarized more compactly by the propensity score. We estimate it with logistic regression using the observed pre-treatment covariates.

```{r}
ps_mod <- glm(
  treat ~ age + educ + black + hispan + married + nodegree + re74 + re75,
  data = aa,
  family = binomial()
)

aa_ps <- ps_mod |>
  augment(data = aa, type.predict = "response") |>
  mutate(ps = .fitted) |>
  select(-.fitted)
```

Now we look at the distributions of the estimated propensity scores in the treated and control groups.

```{r fig.height=4}
ggplot(aa_ps, aes(x = ps, fill = treat)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 20) +
  labs(x = "Estimated propensity score", y = "Count", fill = NULL)
```

The treated units tend to have higher estimated probabilities of treatment than the control units. This is the expected direction. The issue is not that the graph is backwards. The issue is that the groups may be too separated. Good causal adjustment needs overlap, not simply a treatment model that predicts treatment well. If treated and control units have little overlap in propensity score, then there are few comparable units to compare.

The main lesson from this section is that adjusted comparisons only deserve a causal interpretation under assumptions. Consistency tells us that the treatment must be defined clearly enough for the potential outcomes to make sense. Ignorability says that after conditioning on the observed pre-treatment covariates, treatment assignment is as good as random. Positivity says that comparable treated and control units must exist. SUTVA reminds us that one unit's treatment should not directly affect another unit's outcome. Finally, the discussion of mediators, colliders, and bad controls reminds us that adjustment can go wrong when we condition on the wrong variables. The next section uses propensity scores more directly to organize causal comparisons in observational data.
